### 2023-06-27 litrerview_notes_syntheticlethality/paper0904.tex
Add files via upload

\documentclass[10pt]{article}
%\documentclass[twocolumn]{article}
%\documentclass[onecolumn]{article}
% \usepackage{scrtime} % for \thistime (this package MUST be listed first!)
\DeclareUnicodeCharacter{0301}{\'{e}}
\usepackage{times}
\usepackage{graphicx}
\usepackage{float}
\usepackage[margin=0.75in]{geometry}
\usepackage{fancyhdr}
\usepackage{caption}
\usepackage{notoccite}
\usepackage{pgfplotstable}
\usepackage{soul} %For highlights using \hl
%\usepackage[round]{natbib}
%\setcitestyle{aysep={}} %removes the comma between the author and year in citations
%\usepackage{underscore}
\usepackage{pdfpages}
\usepackage{xcolor,colortbl}%for changing cell colour
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{capt-of}
\pagestyle{fancy}
\setlength{\headheight}{15.2pt}
\setlength{\headsep}{13 pt}
\setlength{\parindent}{28 pt}
\setlength{\parskip}{12 pt}
\pagestyle{fancyplain}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
% \usepackage{color,amsmath,amssymb,amsthm,mathrsfs,amsfonts,dsfont}
\usepackage{xspace}
\usepackage{tikz-cd}
\usepackage{tikz}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{calc, arrows}
\usetikzlibrary{external}
\usepackage{pgfplots}
\pgfplotsset{layers/my layer set/.define layer set={background,main,foreground}{},
        set layers=my layer set,}

\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

%for code
\lstdefinestyle{mystyle}{
	backgroundcolor=\color{backcolour},   
	commentstyle=\color{codegreen},
	keywordstyle=\color{magenta},
	numberstyle=\tiny\color{codegray},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	captionpos=b,                    
	keepspaces=true,                 
	numbers=left,                    
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2
}

\lstset{style=mystyle}
% \usetikzlibrary{pgfplots.clickable}
% \usepgfplotslibrary{clickable}
% tables
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{multirow}
% figs
\captionsetup[figure]{labelfont={color=blue}, font={color=black}}
%\usepackage{subfig}% http://ctan.org/pkg/subfig
\usepackage{subcaption}
%\newsubfloat{figure}% Allow sub-figures
\usepackage{caption}%lable fig caption as fig
%\captionsetup[subfigure]{labelfont=bf, justification=raggedright, labelformat=empty} %no caption label
\usepackage{stackengine} %places caption inside figure?
\captionsetup{subrefformat=empty} %when you reference the subcaption it will be (a) for example %{labelfont={color=blue}}
%\captionsetup[subfigure]{labelsep=colon}

% \usepackage{acronym}
% \usepackage{lineno}%for line numbers
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[backend=biber, giveninits=true, doi=false, isbn=false, natbib=true, url=true, eprint=false, style=authoryear-comp, sorting=nyt, sortcites=ynt, maxcitenames=2, maxbibnames=10, minbibnames = 10, uniquename=false, uniquelist=false, dashed=false]{biblatex} % can change the maxbibnames to cut long author lists to specified length followed by et al., currently set to 99.

%% bibliography for each chapter...
\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,thesis,unpublished]{title}{#1\isdot} % removes quotes around title
\renewbibmacro*{volume+number+eid}{%
	\printfield{volume}%
	%  \setunit*{\adddot}% DELETED
	\printfield{number}%
	\setunit{\space}%
	\printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
%\renewcommand*{\newunitpunct}{\space} % remove period after date, but I like it. 
\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}} % this remove the "In: Journal Name" from articles in the bibliography, which happens with the ynt 
\renewbibmacro*{note+pages}{%
	\printfield{note}%
	\setunit{,\space}% could add punctuation here for after volume
	\printfield{pages}%
	\newunit}    
\DefineBibliographyStrings{english}{% clears the pp from pages
	page = {\ifbibliography{}{\adddot}},
	pages = {\ifbibliography{}{\adddot}},
} 
\DeclareFieldFormat{journaltitle}{#1\isdot}
\renewcommand*{\revsdnamepunct}{}%remove comma between last name and first name
\DeclareNameAlias{sortname}{family-given}
% \DeclareNameAlias{sortname}{last-first}
\renewcommand*{\nameyeardelim}{\addspace} % remove comma in text between name and date
\addbibresource{ABC1.bib} % The filename of the bibliography
\usepackage[autostyle=true]{csquotes} % Required to generate language-dependent quotes in the bibliography
\renewrobustcmd*{\bibinitperiod}{}
% you'll have to play with the citation styles to resemble the standard in your field, or just leave them as is here. 
% or, if there is a bst file you like, just get rid of all this biblatex stuff and go back to bibtex. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% generally hyperref needs to be loaded last
\usepackage[hidelinks,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
%\usepackage[hidelinks,colorlinks=false,citecolor=blue,urlcolor=darkbrown]{hyperref}
\tikzexternalize

\lhead{Synthetic Lethality Overview} %This needs to change
\rhead{Alexander Turco}
\title{\sc Building sex-specific synthetic lethality network in cancer - Temp Name}
\author{\sc Alexander Turco}

\providecommand{\figref}[1]{Figure \ref{#1}}  %what?
\providecommand{\tabref}[1]{(Table \ref{#1})}  %what?
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\newcommand{\seg}{\texttt{Seg}\xspace}
\newcommand{\ecoli}{\mbox{\textit{E.\,coli}}\xspace}
\newcommand{\sclong}{\textit{Saccharomyces cerevisiae}\xspace}
\newcommand{\scshrt}{\mbox{\textit{S.\,cerevisiae}}\xspace}
\newcommand{\sce}{\mbox{\textit{S.\,cerevisiae}}\xspace}
\newcommand{\hslong}{\textit{Homo sapiens}\xspace}
\newcommand{\hsshrt}{\mbox{\textit{H.\,sapiens}}\xspace}
\newcommand{\hse}{\mbox{\textit{H.\,sapiens}}\xspace}
\newcommand{\celong}{\textit{Caenorhabditis elegans}\xspace}
\newcommand{\ceshrt}{\mbox{\textit{C.\,elegans}}\xspace}
\newcommand{\dmlong}{\textit{Drosophila melanogaster}\xspace}
\newcommand{\dmshrt}{\mbox{\textit{D.\,melanogaster}}\xspace}
\newcommand{\atlong}{\textit{Arabidopsis thaliana}\xspace}
\newcommand{\atshrt}{\mbox{\textit{A.\,thaliana}}\xspace}
\newcommand{\pflong}{\textit{Plasmodium falciparum}\xspace}
\newcommand{\pfshrt}{\mbox{\textit{P.\,falciparum}}\xspace}
%%%%BIBLIOGRAPHY

%Supplementary File Table Numbers:
\newcommand{\expdata}{S1\xspace}
\newcommand{\seqdata}{S2\xspace}
\newcommand{\protdata}{S3\xspace}
\newcommand{\blast}{S4\xspace}
\newcommand{\tabvar}{S5\xspace}
%Supplementary File Fig Numbers:

\newcommand{\expcor}{S1\xspace}
\newcommand{\expdistATCC}{S3\xspace}
\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\newcommand{\beginsupplement}{%
	\setcounter{table}{0}
	\renewcommand{\thetable}{S\arabic{table}}%    %thetable references table counter 
	\setcounter{figure}{0}
	\renewcommand{\thefigure}{S\arabic{figure}}%
	\setcounter{equation}{0}
	\renewcommand{\theequation}{S\arabic{equation}}%

}
\renewcommand{\thesection}{}
\renewcommand{\thesubsection}{}
\renewcommand{\thesubsubsection}{}
\usepackage{setspace}
%adjust spacing
\doublespacing
\usepackage{titlesec}

\titlespacing\section{0pt}{12pt plus 2pt minus 2pt}{0pt plus 1pt minus 1pt}
\titlespacing\subsection{0pt}{12pt plus 2pt minus 2pt}{0pt plus 1pt minus 1pt}
\titlespacing\subsubsection{0pt}{12pt plus 2pt minus 2pt}{0pt plus 1pt minus 1pt}

% below 3 lines will put ALL table captions at the top...not sure if
% this is what we want but it is good enough for now

% \usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

        \definecolor{atomictangerine}{rgb}{1.0, 0.6, 0.4}
        \definecolor{darkbrown}{rgb}{1.0, 0.56, 0.24}
        \colorlet{darkcol}{black!30!white}
        \colorlet{lightcol}{black!10!white}
        \definecolor{txtcol}{HTML}{F40000}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\widowpenalty10000
\clubpenalty10000

%\linenumbers %for line numbers
\onecolumn
%\twocolumn[  
%       \begin{@twocolumnfalse}
%               \begin{center}
                        \maketitle
%               \end{center}
%                       \bigskip

\thispagestyle{empty}
\noindent \textsuperscript{1} Department of, University, , ON, Canada

\newpage
%\tableofcontents %Will add a nice Table of Contents
\newpage

%\section{Abstract}

%\section{Literature Review/Introduction}

%	\subsection{What are Low Complexity Regions?}

%	\subsection{Characteristics and Types of LCRs}

%	\subsection{Why care about LCRs?}

%	\subsection{How do LCRs Evolve?}

%	\subsection{What is an Approximate Bayesian Computation Markov chain Monte Carlo algorithm?}

%	\subsection{How will we use an ABC-MCMC - (Oct 28 Proposal, To See New Process Refer to}

%\section{Introduction}

%\section{Materials and Methods (mid-year stuff Jan 20)} 
%\label{methods}
%Custom scripts and commands utilized in this analysis can be found on \texttt{GitHub} at
%\url{https://github.com/opticrom/abcmcmc-thesis4c12}.

%\sloppy For a detailed protocol, see Supplementary files on \texttt{GitHub} at
 %\url{https://github.com/JohannaEnright/LCREntropyProject/}.

%	\subsection{ABC-MCMC: The Algorithm}

%	\subsection{Parameters and Summary Statistics}

%	\subsection{Simulation Step: Creation and Mutation of Protein Sequences}

%	\subsection{Normalization and Euclidean Distance Calculation}

%	\subsection{One Sample T-Test for Probability of Acceptance}

%\section{Results}

%	\subsection{Assessing the Evolution Simulator}

%	\subsection{Applying the Simulator to an ABC-MCMC}

%\section{Discussion}

%	\subsection{Mutations Have the Ability to Destroy LCR Formation}

%	\subsection{Does the Length of a Repeat Play a Role in Insertions/Deletions}

%	\subsection{Why Parameters Could not be Estimated}

%	\subsection{Conclusion}

\newpage
\section{Notes}

\subsection{Basic Basic Explanation of Synthetic Lethality}
Definition of synthetic lethality: Describes a situation in which mutations in two genes together result in cell death, but a mutation in either gene alone does not. Cancer cells that only have one mutated gene in a specific pair of genes can depend on the normal partner gene for survival. Interfering with the function of the normal partner gene may cause cancer cells to die. Studying synthetic lethality may help researchers learn more about the function of genes and develop new drugs to treat cancer (taken from \url{https://www.cancer.gov/publications/dictionaries/cancer-terms/def/synthetic-lethality})

\subsection{What will my Project Explore}
\begin{itemize}
	\item There are sex differences in cancer (some more prevalent in males, some in females)
	\item From my understanding (May 9), what we are trying to do is build synthetic lethality networks in sex specific manner (basically can we build these synthetic lethal pairs in a sex-specific manner)
	\item Synthetic lethality is influenced by sex (maybe its a negative interaction, this can rescue cells)
	\item Been told to pay attention to the datasets utilized in the following papers so I know which data I can use for my project (gene expression data, CRISPR knockout datasets, siRNA, etc)
\end{itemize}

\subsection{Types of Data Utilized for Identifying Synthetic Lethal Relationships - How to discover therapeutically relevant SL interactions}
\begin{itemize}
	\item GTEx data: The Genotype-Tissue Expression project provides gene expression data across multiple human tissues - can be utilized to identify tissue specific or context-specific synthetic lethal interactions by comparing gene expression profiles across different tissues or conditions
	\item used in \citep{cheng2021synthetic}
	\item CRISPR/Cas9 Screens - allow researchers to systematically perturb genes and assess their effects on cellular phenotypes - can provide genetic interaction data for identifying synthetic lethal relationships
	\item Current efforts to map SL interactions can be separated into 4 approaches - Statistical approaches which leverage large populations of tumor genomes, evolutionary conservation, Direct experimentation in human cancer cell lines, combinatorial rather than single gene disruptions \citep{shen2018synthetic}
	\item TCGA is a comprehensive database that provides molecular characterization of various cancer types - it provides whole genome sequencing data, transcriptomics data (RNA-seq), epigenomic data, clinical data, etc
	\item DepMap
\end{itemize}

\subsection{questions/confused}
\begin{itemize}
	\item when talking about SL, the literature often says inactivation of either gene (does this inactivation mean its completely removed or does it also entail mutations that could cause a lower expression of the gene?)
	\item I have not heard the literature mention the mechanisms of synthetic lethality - so if both genes are affected, how does the cell actually die or have its viability reduced. Is it just due to the reduced expression of the genes involved?
	\item debMap
	\item X chromosome inactivation - we usually have to infer for females which one is inactivated - infer based on gene expression data etc
	\item cell line I use, know if it comes from male or female -
	\item Do I want to focus on differences in autosomal or sex chromosomes - these are two different questions - 
	\item Why do cancer cells have a tendency to lose or gain sex chromasomes
	\item Sex chromosome dosage??
	\item The biology behind a cell line
\end{itemize}

\subsection{Potential Research Questions}
1. There are sex specific synthetic lethal gene pairs in certain types of cancers (are there gene combinations that result in synthetic lethality in males but not in females, or vice versa)
-

2. Sex specific genetic modifiers affect synthetic lethality in cancer (Are there genes or genetic pathways that act as modifiers of synthetic lethality, specifically in males or females, potentially influencing treatment response or prognosis?)
\newpage

\subsection{PROPOSAL/WORKFLOW}
MAY 18 - Today is my fourth day at UHN, this is the first workflow
\begin{enumerate}
	\item Determine differentially expressed genes between Healthy (non-cancer) and Affected (cancer) Males and Females. Differential gene expression analysis can be utilized to find genes of interest in each of the sexes
	\begin{itemize}
		\item We can obtain cancerous cell line data from CCLE (It has expression data, metadata for the samples, everything necessary for differential expression analysis)
		\item Question: Are there any good databases for healthy cell lines
		\item Question: In the study by \citet{shohat2022gene}, they inferred the presence of sex chromosomes in both male and female cancer cell lines (cancer cells have tendency to lose or gain sex chromosomes) - Should I be taking the same approach for healthy cell lines and infer which chromosomes are present? (Nad mentioned overtime weird things can happen to these cell lines so it could be good to check)
		\item Possibility that there will be much more data on cancer cell lines than healthy cell lines which could affect results - an alternative approach is to directly compare healthy and affected cell lines (for example Healthy breast Tissue and Breast Cancer Tissue) individually as opposed to having many more cancerous cell lines than healthy cell lines and grouping them together.
	\end{itemize}
	\item Identify potential synthetic lethal partners of the genes that were found to be differentially expressed in each sex
	\begin{itemize}
		\item This can be done through utilizing CRISPR-Cas9 knockout data - there is CRISPR knockout data pertaining to cancerous cell lines on CCLE database - have to look in to how to find the interactions (I have seen methods like ISLE but I don't know how relevant it would be here)
		\item Since the genes of interest were differentially expressed in cancer cell lines, does this mean they are potentially contributing to cancer and hence finding specific synthetic lethal partners to these genes could lead to finding sex-specific potential drug targets?
	\end{itemize}
	\item Assess the differences in synthetic lethal partners in male vs females - we may want to exclude genes that we know for sure are only in one sex such as testicular genes - excluding these means we will focus on the differences but it could also be good to keep genes known to be found in one sex as validation/confirmation that they are very highly differentially expressed in each sex (to confirm DGE analysis).
\end{enumerate}

\clearpage\newpage
\section{References}

%%%FIGURES%%%%%

%%%%PRINTING BIBLIOGRAPHY%%%%
\nocite{*}
\printbibliography[heading=none, sorting=nyt]
\newpage

%The following files included these appendices are located at \texttt{/home/alext/scratch/abcmcmc-thesis4c12}

%\section{Appendix 1: Low Complexity Region Evolution Simulator}
%\texttt{/home/alext/scratch/abcmcmc-thesis4c12/mutations\_2\_BG\_vecs.cpp}
%\lstinputlisting[language=c++]{../mutations_2_BG_vecs.cpp}
%\newpage

%\section{Appendix 2: ABC-MCMC in C++}
%\texttt{/home/alext/scratch/abcmcmc-thesis4c12/abcmcmc.cpp}
%\lstinputlisting[language=c++]{../abcmcmc.cpp}
%\newpage

%\section{Appendix 3: Simulated Protein Summary Statistics}
%\texttt{/home/alext/scratch/abcmcmc-thesis4c12/abcmcmc.cpp/simulated\_protein.cpp}
%\lstinputlisting[language=c++]{../simulated_protein.cpp}
%\newpage

%\section{Appendix 4: Euclidean Distance and Normlization of Vectors}
%\texttt{/home/alext/scratch/abcmcmc-thesis4c12/abcmcmc.cpp/distance2.cpp}
%\lstinputlisting[language=c++]{../distance2.cpp}
%\newpage

%\section{Appendix 5: One Sample t-test}
%\texttt{/home/alext/scratch/abcmcmc-thesis4c12/abcmcmc.cpp/ttest.cpp}
%\lstinputlisting[language=c++]{../ttest.cpp}
%\newpage

\end{document}

### 2023-06-27 sl_report_pmcrc/paper0904.tex
updates

--- 2023-06-27 litrerview_notes_syntheticlethality/paper0904.tex
@@
 
-\lhead{Synthetic Lethality Overview} %This needs to change
+\lhead{Synthetic Lethality} %This needs to change
 \rhead{Alexander Turco}
@@
 \newpage
-%\tableofcontents %Will add a nice Table of Contents
+\tableofcontents %Will add a nice Table of Contents
 \newpage
 
-%\section{Abstract}
+\section{Abstract}
 
-%\section{Literature Review/Introduction}
+\newpage
+\section{Introduction/Background Information}
 
-%	\subsection{What are Low Complexity Regions?}
+	%What are they
+	\subsection{What are Synthetic Lethal Interactions?}
 
-%	\subsection{Characteristics and Types of LCRs}
+	%Why are they important
+	\subsection{Synthetic Lethal Interactions are Harnessed for Precision Oncology}
 
-%	\subsection{Why care about LCRs?}
+	%Building SL networks Pan-cancer
+	\subsection{Building Pan-Cancer Synthetic Lethality Networks}
 
-%	\subsection{How do LCRs Evolve?}
+	%Bringing sex differences into the paper
+	\subsection{Human Sex Differences add An Additional Layer of Complexity}
 
-%	\subsection{What is an Approximate Bayesian Computation Markov chain Monte Carlo algorithm?}
+	%Bringing everything together
+	\subsection{Building Pan-Cancer Synthetic Lethality Networks in a Sex Specific Manner}
 
-%	\subsection{How will we use an ABC-MCMC - (Oct 28 Proposal, To See New Process Refer to}
-
-%\section{Introduction}
-
-%\section{Materials and Methods (mid-year stuff Jan 20)} 
+\newpage
+\section{Materials and Methods} 
 %\label{methods}
-%Custom scripts and commands utilized in this analysis can be found on \texttt{GitHub} at
-%\url{https://github.com/opticrom/abcmcmc-thesis4c12}.
+%Custom scripts and commands utilized in this analysis can be found on \texttt{GitHub} at \url{https://github.com/opticrom/abcmcmc-thesis4c12}.
 
@@
 
-%	\subsection{ABC-MCMC: The Algorithm}
+	\subsection{TCGA Data}
 
-%	\subsection{Parameters and Summary Statistics}
+	\subsection{Pre-filtering and Normalization of Raw RNA-seq Count Data}
 
-%	\subsection{Simulation Step: Creation and Mutation of Protein Sequences}
+	\subsection{Data Quality Assessment (PCA, NPManova)}
 
-%	\subsection{Normalization and Euclidean Distance Calculation}
+	\subsection{Differential Gene Expression Analysis with DESeq2}
 
@@
 
-%\section{Results}
+\section{Results}
 
@@
 
-%\section{Discussion}
+\section{Discussion}
 
@@
 %	\subsection{Conclusion}
-
-\newpage
-\section{Notes}
-
-\subsection{Basic Basic Explanation of Synthetic Lethality}
-Definition of synthetic lethality: Describes a situation in which mutations in two genes together result in cell death, but a mutation in either gene alone does not.
-Cancer cells that only have one mutated gene in a specific pair of genes can depend on the normal partner gene for survival.
-Interfering with the function of the normal partner gene may cause cancer cells to die.
-Studying synthetic lethality may help researchers learn more about the function of genes and develop new drugs to treat cancer (taken from \url{https://www.cancer.gov/publications/dictionaries/cancer-terms/def/synthetic-lethality})
-
-\subsection{What will my Project Explore}
-\begin{itemize}
-	\item There are sex differences in cancer (some more prevalent in males, some in females)
-	\item From my understanding (May 9), what we are trying to do is build synthetic lethality networks in sex specific manner (basically can we build these synthetic lethal pairs in a sex-specific manner)
-	\item Synthetic lethality is influenced by sex (maybe its a negative interaction, this can rescue cells)
-	\item Been told to pay attention to the datasets utilized in the following papers so I know which data I can use for my project (gene expression data, CRISPR knockout datasets, siRNA, etc)
-\end{itemize}
-
-\subsection{Types of Data Utilized for Identifying Synthetic Lethal Relationships - How to discover therapeutically relevant SL interactions}
-\begin{itemize}
-	\item GTEx data: The Genotype-Tissue Expression project provides gene expression data across multiple human tissues - can be utilized to identify tissue specific or context-specific synthetic lethal interactions by comparing gene expression profiles across different tissues or conditions
-	\item used in \citep{cheng2021synthetic}
-	\item CRISPR/Cas9 Screens - allow researchers to systematically perturb genes and assess their effects on cellular phenotypes - can provide genetic interaction data for identifying synthetic lethal relationships
-	\item Current efforts to map SL interactions can be separated into 4 approaches - Statistical approaches which leverage large populations of tumor genomes, evolutionary conservation, Direct experimentation in human cancer cell lines, combinatorial rather than single gene disruptions \citep{shen2018synthetic}
-	\item TCGA is a comprehensive database that provides molecular characterization of various cancer types - it provides whole genome sequencing data, transcriptomics data (RNA-seq), epigenomic data, clinical data, etc
-	\item DepMap
-\end{itemize}
-
-\subsection{questions/confused}
-\begin{itemize}
-	\item when talking about SL, the literature often says inactivation of either gene (does this inactivation mean its completely removed or does it also entail mutations that could cause a lower expression of the gene?)
-	\item I have not heard the literature mention the mechanisms of synthetic lethality - so if both genes are affected, how does the cell actually die or have its viability reduced.
-Is it just due to the reduced expression of the genes involved?
-	\item debMap
-	\item X chromosome inactivation - we usually have to infer for females which one is inactivated - infer based on gene expression data etc
-	\item cell line I use, know if it comes from male or female -
-	\item Do I want to focus on differences in autosomal or sex chromosomes - these are two different questions - 
-	\item Why do cancer cells have a tendency to lose or gain sex chromasomes
-	\item Sex chromosome dosage??
-	\item The biology behind a cell line
-\end{itemize}
-
-\subsection{Potential Research Questions}
-1. There are sex specific synthetic lethal gene pairs in certain types of cancers (are there gene combinations that result in synthetic lethality in males but not in females, or vice versa)
--
-
-2.
-Sex specific genetic modifiers affect synthetic lethality in cancer (Are there genes or genetic pathways that act as modifiers of synthetic lethality, specifically in males or females, potentially influencing treatment response or prognosis?)
-\newpage
-
-\subsection{PROPOSAL/WORKFLOW}
-MAY 18 - Today is my fourth day at UHN, this is the first workflow
-\begin{enumerate}
-	\item Determine differentially expressed genes between Healthy (non-cancer) and Affected (cancer) Males and Females.
-Differential gene expression analysis can be utilized to find genes of interest in each of the sexes
-	\begin{itemize}
-		\item We can obtain cancerous cell line data from CCLE (It has expression data, metadata for the samples, everything necessary for differential expression analysis)
-		\item Question: Are there any good databases for healthy cell lines
-		\item Question: In the study by \citet{shohat2022gene}, they inferred the presence of sex chromosomes in both male and female cancer cell lines (cancer cells have tendency to lose or gain sex chromosomes) - Should I be taking the same approach for healthy cell lines and infer which chromosomes are present?
-(Nad mentioned overtime weird things can happen to these cell lines so it could be good to check)
-		\item Possibility that there will be much more data on cancer cell lines than healthy cell lines which could affect results - an alternative approach is to directly compare healthy and affected cell lines (for example Healthy breast Tissue and Breast Cancer Tissue) individually as opposed to having many more cancerous cell lines than healthy cell lines and grouping them together.
-	\end{itemize}
-	\item Identify potential synthetic lethal partners of the genes that were found to be differentially expressed in each sex
-	\begin{itemize}
-		\item This can be done through utilizing CRISPR-Cas9 knockout data - there is CRISPR knockout data pertaining to cancerous cell lines on CCLE database - have to look in to how to find the interactions (I have seen methods like ISLE but I don't know how relevant it would be here)
-		\item Since the genes of interest were differentially expressed in cancer cell lines, does this mean they are potentially contributing to cancer and hence finding specific synthetic lethal partners to these genes could lead to finding sex-specific potential drug targets?
-	\end{itemize}
-	\item Assess the differences in synthetic lethal partners in male vs females - we may want to exclude genes that we know for sure are only in one sex such as testicular genes - excluding these means we will focus on the differences but it could also be good to keep genes known to be found in one sex as validation/confirmation that they are very highly differentially expressed in each sex (to confirm DGE analysis).
-\end{enumerate}
 

### 2023-06-27 sl_report_pmcrc/mutations_2.cpp

@@ -1,0 +1,170 @@
+#include "functions.cpp"
+#include "getindex.cpp"
+#include <bits/stdc++.h>
+#include <iostream>
+#include <vector>
+#include <string>
+#include <ctime>
+#include <algorithm> //this is to get min element stuff, cool library
+#define numAA 20
+using namespace std;
+
+
+Ran myran(time(NULL)); //We will use 21 as the random seed right now, used in Poissondev too
+
+////////////////////////////////////////////////////////////////
+// FIRST FUNCTION TAKES AN INTEGER VALUE AND GENERATES A RANDOM/
+// AMINO ACID SEQUENCE OF THAT LENGTH                          / 
+////////////////////////////////////////////////////////////////
+
+std::string createSeq(int n){
+
+    char aminoAcids[numAA] = { 'G', 'A', 'L', 'M', 'F', 'W', 'K', 'Q', 'E', 'S', 'P', 'V', 
+                                   'I', 'C', 'Y', 'H', 'R', 'N', 'D', 'T' };
+
+    std::string protein = "";
+    for (int i = 0; i < n; i++){
+        protein += aminoAcids[myran.int64() % numAA];} //this rand() % 20 means in the range 0-19
+    
+    //std::cout << protein << "\n" << "\n" ;
+    return protein;
+}
+
+
+/*THIS FUNCTION WILL MUTATE THE SIMULATED PROTEIN SEQUENCE
+BY CHOOSING A RANDOM EXPONENTIAL DEVIATE (WITH MEAN = MUTATION RATE)
+FOR EACH AMINO ACID IN THE SEQUENCE AND SUBSEQUENTLY SELECTING
+THE AMINO ACID WITH THE LOWEST NUMBER (QUICKEST TO MUTATE) AND
+MUTATING IT RANDOMLY, THIS IS DONE SUCCESSIVELY TO PRODUCE A 
+PROTEIN AND CREATE A VECTOR OF VALUES SIMILAR TO ABOVE (SEP 21)     
+SEP 26 - PROCESS CHANGE, THIS FUNCTION WILL BE USED FOR AMINO ACID EXPANSION
+UPDATE NOV 15 - WE ARE MAKING THIS NEW FUNCTION THAT ASSIGNS DEVIATES
+BASED ON BOTH MUTATION RATES AND INDEL RATES AND THEN OUT OF BOTH VECTORS
+WE WILL FIND THE LOWEST AND EITHER MUTATE IT OR DO AN INS/DEL DEPENDING
[…]

### 2023-07-10 july18meeting_pres2/may30ccgmeeting.tex
adding second presentation latex

% Inbuilt themes in beamer
\documentclass{beamer}

%Defining some colours
\definecolor{darkred}{rgb}{0.8,0,0}

% Theme choice: there a number of preset themes to choose from
% Play around with them, Cambridge is nice for first pres
%\usetheme{Szeged}
\usetheme{CambridgeUS} %setting the main theme
\usecolortheme{whale} % setting colour theme
\usefonttheme{professionalfonts} %font theme
\useinnertheme[shadow=true]{rounded}
%\useoutertheme{} %outer theme

%\setbeamertemplate{footline} %Remove footer line in all slides
\setbeamertemplate{navigation symbols}{} %removes navigation symbols
\setbeamertemplate{footline}[page number] %removes footer line, keeps pg#
\setbeamertemplate{caption}{\insertcaption}

%Setting colours for boxes and captions
\setbeamercolor{block title}{bg=blue!30, fg=black}
\setbeamercolor{block body}{bg=blue!10}
\setbeamercolor{frametitle}{fg=black}

%%%FOR VIDEOS
\usepackage{media9}
\usepackage{multimedia}
% For the Flowchart
%\usepackage{tikz}
%\usetikzlibrary{shapes.geometric, arrows}
%\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30]
%\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
%\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=green!30]
%\tikzstyle{arrow} = [thick,->,>=stealth]

%%% For the Timeline
\usepackage{xcolor}
\usepackage{tikz} \usetikzlibrary{calc, arrows.meta, intersections, patterns, positioning, shapes.misc, fadings, through,decorations.pathreplacing}

\definecolor{ColorOne}{rgb}{0.0,0.5,1.0} %Lightblue
\definecolor{ColorTwo}{rgb}{1.0,0.6,0.4} %lightorange
\definecolor{ColorThree}{rgb}{0.75,0.58,0.89} %lightpurple

% Title page details: 
\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers} 
\author{Alexander Turco}
\date{\today}
\logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}

% Bibliography stuff
\usepackage[natbib=true, sorting=nyt, style=authoryear-comp]{biblatex}
\addbibresource{ABC1.bib}

%Extra packages
\usepackage{makecell}

%For itemize
\setbeamertemplate{itemize item}[triangle]

%For Flow Chart
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

\tikzstyle{startstop} = [rectangle, 
minimum width=6cm, 
minimum height=1cm,
text centered, 
draw=black, 
fill=blue!30]

\tikzstyle{io} = [trapezium, 
trapezium stretches=true, % A later addition
trapezium left angle=70, 
trapezium right angle=110, 
minimum width=3cm, 
minimum height=1cm, text centered, 
draw=black, fill=blue!30]

\tikzstyle{process} = [rectangle, 
minimum width=3cm, 
minimum height=1cm, 
text align, 
text width=1cm, 
draw=black, 
fill=orange!30]

\tikzstyle{decision} = [diamond, 
minimum width=3cm, 
minimum height=1cm, 
text centered, 
draw=black, 
fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\begin{document}

	% For my introduction slides, there will be the slide with my title and name as well as an outline slide with the brief overview of what I will discuss.
	% Title page frame - SLIDE 1%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\section{Introduction}
	\begin{frame}
		\titlepage 
		\begin{center}
			%\includegraphics[width=7cm, height=4cm]{cellsatwar.png}
		\end{center}
	\end{frame}

	% Remove logo from the next slides
	\logo{}

	% Outline frame - SLIDE 2
	%\begin{frame}{Overview}

		%\begin{center}
		%\begin{minipage}{6cm}

		 % 		\begin{block}{} \hyperlink{link1}{Background Information} \end{block}
		  %		\begin{block}{} The Game Design Process \end{block}
		  %		\begin{block}{} Collection of Student Feedback \end{block}
		  %		\begin{block}{} Results \end{block}
		  		%\begin{block}{} Conclusions \end{block}
		  %		\begin{block}{} Future Work \end{block}

		%\end{minipage}
		%\end{center}

	%\end{frame}

	% For my Background slides, I will talk about important things such as WHat LCRs are, their evolution, stuff like that
	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
	\section{Background}
	\begin{frame}{Synthetic Lethality has Been Around for Some Time}
		\begin{itemize}
			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
		\end{itemize}

		\centering
		\huge \textbf{Synthetic = Combining}

		\footnotetext[1]{\tiny\cite{nijman2011synthetic}}

	\end{frame}

	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
		\label{link1}

		\begin{itemize}
			\item Synthetic lethal interactions describe the relationship between two genes whose coupled inactivation, but not their individual inactivation, causes cell death or reduces cell viability \newline

			\includegraphics[width=11cm, height=2.5cm]{slbasic.png}

			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
		\end{itemize}

	\end{frame}

	% SLIDE 4 - I do Not Know if I am including this slide
	% The Graphic for this slide is an  
	% Entropy, which is measured by Shannon's Entropy equation is a measure of compositional complexity which uses the proportion of residue(s)
	% in a subsequence to measure the compositional state of that subseequence - A lower variety of residues = lower entropy 
	%\begin{frame}{Shannon's Entropy - MAYBE }

	%		$H = -L\sum p_i log_2(p_i)$

	%\end{frame}

	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}

		\begin{block}{\textit{To gain insight into gene function}}
			 \begin{itemize}
			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
			 \end{itemize}
		\end{block} \pause

		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
			\begin{itemize}
				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g. Olaparib was the first drug to work via a synthetic lethal mechanism)
				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
				%Talk about BRCA1 and PARP inhibitors for olaparib lol
			\end{itemize}
		\end{block}

	\footnotetext[3]{\tiny\cite{nijman2011synthetic,shen2018synthetic}}

	\end{frame}

	%Slide 6
	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway. PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process. In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
		\begin{itemize}
			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
		\end{itemize}
		\centering
		\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}

		\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
	\end{frame}

	%SLIDE 6 
	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}

	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics. Some aspects of this phenomena are; \newline

	\begin{enumerate}
		\item Variation in lifetime cancer risk
		\item Variation in cancer onset age
		\item Variation in the genes driving the cancer across tissue types \newline
	\end{enumerate}
		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
	\end{frame}

	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
			\begin{center}
				\includegraphics[width=10cm, height=5.7cm]{csl.png}
			\end{center}
		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
	\end{frame}

	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
	\section{Approaches}

	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type

		\centering
		\includegraphics[width=8cm, height=6cm]{tcl1.png}
		\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}

		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
		\begin{columns}
			\begin{column}{0.5\textwidth}
					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
					\centering
					Stem Cell Divisions
			\end{column}
			\begin{column}{0.5\textwidth}
				\includegraphics[width=6cm, height=3cm]{tcl3.png}
				\centering
				\vspace{-0.5cm}Methylation
			\end{column}
		\end{columns}	
		\centering	
	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}

	\begin{frame}{Where did the Data come From?}
		A recently published reference set of genome-wide cSLs common to many cancer types was utilized
		\begin{center}
				\includegraphics[width=6cm,height=6cm]{corenet.png}
		\end{center}
	\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

	%EXPLAIN THE PROCESS OF ISLE HERE
	\begin{frame}{Building Pan-Cancer Synthetic Lethality Networks}
		\centering
		\includegraphics[width=13cm, height=7cm]{isle.png}
		%\movie[width=10cm,height=7cm,showcontrols=true]{\includegraphics[width=10cm, height=7cm]{pompe_thumbnail.png}}{pompe_trimmed.mp4}
		\footnotetext[2]{\tiny\cite{lee2018harnessing}}
	\end{frame}

	%%% EXPLAIN NSSE SURVEY A BIT
	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
	\begin{frame}{There is Variability in Synthetic Lethal Interactions Depending on Cancer Type}

		\textbf{\color{blue}Pan-Cancer Analysis:} Analysis of multiple types of cancer simultaneously, aims to identify commonalities shared across different types

		\textbf{\color{blue}Cancer-Specific Analysis:} Analysis of individual cancer types to better understand the variation within a specific cancer type \vspace{1cm}

		\begin{itemize}
			\item Pan-cancer analyses may overlook cancer-specific differences
			\item Availability of datasets for all cancer types varies, this may limit statistical power of pan-cancer analyses
		\end{itemize}

	\end{frame}

	%%%%RESULTS BEGIN HERE
	\section{Results}

	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
		Human sex differences are mainly caused by;
		\begin{enumerate}
			\item Gonadal hormone secretions
			\item Genes located on the sex chromosomes (X and Y) \newline
		\end{enumerate}
	This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
	\end{frame}

	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
		\centering
		\includegraphics[width=6cm,height=7cm]{sex.png}
		\footnotetext[6]{\tiny\cite{oliva2020impact}}
	\end{frame}

	\begin{frame}{The Objective}
		\begin{center}
		%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
		Can we build sex-specific synthetic lethality networks for various cancer types? \newline

		More specifically, we are trying to elucidate the differences in synthetic lethal interactions between males and females using a network based approach.
		\end{center}
	\end{frame}

\end{document}

### 2023-07-10 scripts_and_executables_examples/adding_Blatmerges_230418.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python3
+
+import pandas as pd
+import glob
+import numpy as np
+from sys import argv
+import os
+import datetime
+from pytz import timezone
+
+print('\n**********************')
+print('START TIME:', datetime.datetime.now(timezone('EST')))
+
+
+# loading in files! --> We're not going to use the scores yet!! --> VERY DIFFERENT, WE'RE JUST COUNTING UP THE NUMBER SO FAR
+res = pd.read_csv(str(argv[1]), sep='\t', header=None)
+print(res.head())
+res.columns = ['matches', 'misMatches', 'repMatches', 'nCount', 'qNumInsert', 'qBaseInsert', 'tNumInsert', 'tBaseInsert', 'strand', 'qName', 'qSize', 'qStart', 'qEnd', 'tName', 'tSize', 'tStart', 'tEnd', 'blockCount', 'blockSizes','qStarts', 'tStarts']
+# sorting the blat responses
+sorted_res = res.sort_values(by=['matches'], ascending=False)
+sorted_res = sorted_res.drop_duplicates(subset='qName', keep="first")
+
+
+# recall, no score csv --> using the 2nd argument for the svs
+svs = pd.read_csv(str(argv[2]), comment='#', sep='\t')
+print('PRE ANNOTATION')
+print(svs.head())
+
+# BY CHROMOSOME!!
+chromo = str(argv[3])
+svs = svs[svs.CHROM == chromo]
+ogrow = len(svs)
+print('og number of rows:', ogrow)
+
+#adding a column for a merge..
+for idx, row in svs.iterrows():
+	svs.loc[idx, 'qName'] = ('POST_'+ row['CHROM']+ '_' +str(row['POS'])+ '_' + str(row['SVlen'])+ '_' + row['SV_Type'])
+
+merged2 = pd.merge(svs, sorted_res, how = 'outer', on = 'qName', indicator = True)
+
+for idx, row in merged2.iterrows():
+	if row['_merge'] == 'left_only':
+		merged2.loc[idx, 'Homology'] = False
+	elif row['_merge'] == 'both':
+		merged2.loc[idx, 'Homology'] = True
+
+newrow = len(merged2)
+print('new number of rows:', newrow)
+
+# checking if the rows match (they should)
+if ogrow == newrow:
+	print('good merge')
[…]

### 2023-07-10 scripts_and_executables_examples/execute_Blatmerges_230418.sh

@@ -1,0 +1,41 @@
+#!/bin/bash
+#SBATCH --nodes=1
+#SBATCH --mem=300000
+#SBATCH --time=08:00:00
+#SBATCH --mail-type=ALL
+#SBATCH --mail-user=nb.boev@mail.utoronto.ca
+#SBATCH --job-name=blatmerge
+#SBATCH --output=blatmerge.o
+#SBATCH --error=blatmerge.e
+#SBATCH --qos=privileged
+
+source $HOME/virtenv/bin/activate
+module load scipy-stack
+
+# starting with the X chromo = 2023/04/18
+python adding_Blatmerges_230418.py \
+/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrX.Blatresults_full.csv \
+/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/SVlen/variants_freeze4_sv_insdel_alt.vcf.SVlength.csv chrX \
+>> ./sv/adding_Blatmerges_230418.py.txt
+echo "Counting for Chromo X"
+rm /home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrX.Blatresults_full.csv
+
+
+# Running on Y chromo
+python adding_Blatmerges_230418.py \
+/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrY.Blatresults_full.csv \
+/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/SVlen/variants_freeze4_sv_insdel_alt.vcf.SVlength.csv chrY \
+>> ./sv/adding_Blatmerges_230418.py.txt
+echo "Counting for Chromo Y"
+rm /home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrY.Blatresults_full.csv
+
+# runnning through autosomes
+for i in {1..22};
+do
+	python adding_Blatmerges_230418.py \
+/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chr${i}.Blatresults_full.csv \
[…]

### 2023-07-10 july18meeting_pres2/may30ccgmeeting.tex
started July 18 presentation

@@
 % Title page details: 
-\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers} 
+\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers \\~\\ Project Updates} 
 \author{Alexander Turco}
-\date{\today}
+\date{July 18, 2023}
 \logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}
@@
 	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
-	\section{Background}
-	\begin{frame}{Synthetic Lethality has Been Around for Some Time}
-		\begin{itemize}
-			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
-			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
-		\end{itemize}
-
-		\centering
-		\huge \textbf{Synthetic = Combining}
-
-		\footnotetext[1]{\tiny\cite{nijman2011synthetic}}
-
-	\end{frame}
 
 	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
-	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
+	\begin{frame}{Recall: The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
 		\label{link1}
@@
 			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
+			\end{itemize}
+			\centering
+			\large \textbf{Synthetic = Combining}
 			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
-		\end{itemize}
-
 	\end{frame}
@@
 
-	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
-	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}
-
-		\begin{block}{\textit{To gain insight into gene function}}
-			 \begin{itemize}
-			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
-			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
-			 \end{itemize}
-		\end{block} \pause
-
-		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
-			\begin{itemize}
-				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g.
-Olaparib was the first drug to work via a synthetic lethal mechanism)
-				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
-				%Talk about BRCA1 and PARP inhibitors for olaparib lol
-			\end{itemize}
-		\end{block}
-
-	\footnotetext[3]{\tiny\cite{nijman2011synthetic,shen2018synthetic}}
-
-	\end{frame}
-
 	%Slide 6
@@
 In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
-	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
-		\begin{itemize}
-			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
-		\end{itemize}
+	\begin{frame}{Recall: Synthetic Lethal Interactions are Harnessed for Precision Oncology}
+		\begin{itemize} %(Olaparib, Niraparib, Rucaparib, Talazoparib)
+			\item \small Four FDA approved anti-cancer drugs  are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors that work via a synthetic lethal mechanism
+
 			\centering
@@
 			\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
-	\end{frame}
+		\end{itemize}
 
-	%SLIDE 6 
-	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}
-
-	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics.
-Some aspects of this phenomena are; \newline
-
-	\begin{enumerate}
-		\item Variation in lifetime cancer risk
-		\item Variation in cancer onset age
-		\item Variation in the genes driving the cancer across tissue types \newline
-	\end{enumerate}
-		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
-	\end{frame}
-
-	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
-			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
-			\begin{center}
-				\includegraphics[width=10cm, height=5.7cm]{csl.png}
-			\end{center}
-		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
-	\end{frame}
-
-	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
-	\section{Approaches}
-
-	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
-		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type
-
-		\centering
-		\includegraphics[width=8cm, height=6cm]{tcl1.png}
-		\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
-	\end{frame}
-
-		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
-		\begin{columns}
-			\begin{column}{0.5\textwidth}
-					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
-					\centering
-					Stem Cell Divisions
-			\end{column}
-			\begin{column}{0.5\textwidth}
-				\includegraphics[width=6cm, height=3cm]{tcl3.png}
-				\centering
-				\vspace{-0.5cm}Methylation
-			\end{column}
-		\end{columns}	
-		\centering	
-	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
-	\end{frame}
-
-	\begin{frame}{Where did the Data come From?}
-		A recently published reference set of genome-wide cSLs common to many cancer types was utilized
-		\begin{center}
-				\includegraphics[width=6cm,height=6cm]{corenet.png}
-		\end{center}
-	\footnotetext[2]{\tiny\cite{lee2018harnessing}}
 	\end{frame}
@@
 	%EXPLAIN THE PROCESS OF ISLE HERE
-	\begin{frame}{Building Pan-Cancer Synthetic Lethality Networks}
+	\begin{frame}{Recall: Building Pan-Cancer Synthetic Lethality Networks}
 		\centering
@@
 
-	%%% EXPLAIN NSSE SURVEY A BIT
-	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
-	\begin{frame}{There is Variability in Synthetic Lethal Interactions Depending on Cancer Type}
-
-		\textbf{\color{blue}Pan-Cancer Analysis:} Analysis of multiple types of cancer simultaneously, aims to identify commonalities shared across different types
-
-		\textbf{\color{blue}Cancer-Specific Analysis:} Analysis of individual cancer types to better understand the variation within a specific cancer type \vspace{1cm}
-
-		\begin{itemize}
-			\item Pan-cancer analyses may overlook cancer-specific differences
-			\item Availability of datasets for all cancer types varies, this may limit statistical power of pan-cancer analyses
-		\end{itemize}
-
-	\end{frame}
-
-	%%%%RESULTS BEGIN HERE
-	\section{Results}
-
-	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
+	\begin{frame}{Recall: Sex Differences Add an Additional Layer of Complexity}
 		Human sex differences are mainly caused by;
@@
 		This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
-	\end{frame}
-
-	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
-		\centering
-		\includegraphics[width=6cm,height=7cm]{sex.png}
-		\footnotetext[6]{\tiny\cite{oliva2020impact}}
 	\end{frame}
@@
 
+	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
+	\section{Materials \& Methods}
+
+	\begin{frame}{Overall Project Workflow}
+		content...
+	\end{frame}
+
+	\begin{frame}{Selection of TCGA Cancer Types}
+		content...
+	\end{frame}
+
+	%%%%RESULTS BEGIN HERE
+	\section{Results}
+
 \end{document}

### 2023-07-10 july18meeting_pres2/may30ccgmeeting.tex
updares

@@
 	\begin{frame}{Overall Project Workflow}
+		\includegraphics[width=12cm, height=5.2cm]{img2.png}
+	\end{frame}
+
+	\begin{frame}{Selection of TCGA Cancer Types}
+			\begin{columns}
+				\begin{column}{0.45\textwidth}
+					\includegraphics[width=5cm, height=8cm]{img1.png}
+				\end{column}
+				\begin{column}{0.5\textwidth}
+					\begin{itemize}
+						\item There is a lack of healthy tissue RNA-seq samples in the TCGA database (12 cancers with more than 10 pairs in M and F)
+					\end{itemize}
+				\end{column}
+			\end{columns}
+
+		\includegraphics[width=5cm, height=8cm]{img1.png}
+	\end{frame}
+
+	\begin{frame}{Creating Pan-Cancer Expression Matrices}
 		content...
@@
 
-	\begin{frame}{Selection of TCGA Cancer Types}
+	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
 		content...

### 2023-07-11 sl_report_pmcrc/paper0904.tex
updates to report

@@
 
+	RNA sequencing (RNA-seq) data was obtained from The Cancer Genome Atlas (TCGA).
+Raw STAR aligned counts for tumor tissue and healthy tissue samples were utilized.
+The Cancer Genome Atlas contains genomic information which spans 33 cancer types.
+For the purpose of this study, only 12 of the 33 cancer types were considered.
+We first filtered out sex-biased cancers which include breast invasive carcinoma (BRCA), cervical cell carcinoma (CESC), ovarian serous cystadenocarcinoma (OV), prostate adenocarcinoma (PRAD), testicular germ cell tumors (TGCT), uterine corpus endometrial carcinoma (UCEC), and uterine carcinosarcoma (UCS).
+Next, we filtered out blood cancers as well as cancers which lacked normal tissue gene expression samples.
+This included adrenocortical carcinoma (ACC), lymphoid neoplasm diffuse large b-cell lymphoma (DLBC), glioblastoma multiforme (GBM), acute myeloid leukemia (LAML), and brain lower grade glioma (LGG).
+Finally, we filtered out cancers that did not have any matching pairs of samples (NT and TP from same individual), as well as cancers that had extremely low numbers of matched sample pairs.
+This included mesothelioma (MESO), skin cutaneous melanoma (SKCM), thymoma (THYM), uveal melanoma (UVM), cholangiocarcinoma (CHOL), pancreatic adenocarcinoma (PAAD), pheochromocytoma and paraganglioma (PCPG), rectum adenocarcinoma (READ), and sarcoma (SARC).
+
 	\subsection{Pre-filtering and Normalization of Raw RNA-seq Count Data}

### 2023-07-12 july18meeting_pres2/may30ccgmeeting.tex
cant add .tiff to latex

@@
 				\begin{column}{0.45\textwidth}
-					\includegraphics[width=5cm, height=8cm]{img1.png}
+					\includegraphics[width=5cm, height=8cm]{img3.png}
 				\end{column}
@@
 						\item There is a lack of healthy tissue RNA-seq samples in the TCGA database (12 cancers with more than 10 pairs in M and F)
+						\item Raw count pan-cancer matrices created with 12 remaining TCGA cancer types (BLCA, COAD, ESCA, HNSC, KICH, KIRC, KIRP, LIHC, LUAD, LUSC, STAD, THCA)
 					\end{itemize}
@@
 			\end{columns}			
-
-		\includegraphics[width=5cm, height=8cm]{img1.png}
 	\end{frame}
 
-	\begin{frame}{Creating Pan-Cancer Expression Matrices}
-		content...
+	\begin{frame}{Creating and Pre-Filtering Pan-Cancer Expression Matrices}
+		\begin{itemize}
+			\item For each gene, if expression is $>$ 90th quantile of overall expression in AT LEAST 1 sample, we keep it, otherwise we filter it out
+		\end{itemize}
 	\end{frame}
@@
 	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
-		content...
+		\includegraphics{all_cancers-M_before_and_after_normalization_v.tiff}
 	\end{frame}

### 2023-07-12 july18meeting_pres2/may30ccgmeeting.tex
updates

@@
 \usepackage{multimedia}
+\usepackage{graphicx}
 % For the Flowchart
@@
 			\item For each gene, if expression is $>$ 90th quantile of overall expression in AT LEAST 1 sample, we keep it, otherwise we filter it out
+			\item We reduce the amount of genes with extremely low expression, thereby reducing noise and improving sensitivity to detect differentially expressed genes (genes with weak expression are more susceptible to technical noise arising from library size, library composition, etc)
 		\end{itemize}
@@
 
+	%Both samples are sequenced at the same depth - sample A has 5 genes expressed at same level, Sample B has the same genes as sample A plus two highly expressed genes that take up 50% of the reads
+	\begin{frame}{Normalization for RNA-seq Data Analysis}
+		\centering Required to identify genes that are differentially expressed due to some biological phenomena and not due to technical variation
+	\end{frame}
+
 	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
-		\includegraphics{all_cancers-M_before_and_after_normalization_v.tiff}
+		\includegraphics[width=12cm, height=8cm]{all_cancers-F_before_and_after_normalization_v.png}
+	\end{frame}
+
+	\begin{frame}
+		content...
 	\end{frame}

### 2023-07-13 july18meeting_pres2/may30ccgmeeting.tex
ppt updates

@@
 
+	\section{Approach}
 	\begin{frame}{The Objective}
@@
 	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
-	\section{Materials \& Methods}
 
@@
 
+	\section{Data Collection}
 	\begin{frame}{Selection of TCGA Cancer Types}
@@
 	%Both samples are sequenced at the same depth - sample A has 5 genes expressed at same level, Sample B has the same genes as sample A plus two highly expressed genes that take up 50% of the reads
+	\section{Data Processing}
 	\begin{frame}{Normalization for RNA-seq Data Analysis}
 		\centering Required to identify genes that are differentially expressed due to some biological phenomena and not due to technical variation
+
+		\includegraphics[width=9cm,height=7cm]{img4.png}
 	\end{frame}
@@
 
-	\begin{frame}
-		content...
+	\section{Data Processing}
+	\begin{frame}{Normal and Tumor Tissue Differ ACROSS 12 Cancer Types}
+		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_0.9.png}
+
+		Across all 12 cancer types, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (NPManova p-val = 0.0001)
 	\end{frame}
 
-	%%%%RESULTS BEGIN HERE
-	\section{Results}
+	\begin{frame}{Normal and Tumor Tissue Differ WITHIN each Cancer}
+		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_cancertype_0.9.png}
+
+		Within each cancer type, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (Acceptions: Males: ESCA (LOW SAMPLE SIZE), Females: BLCA, ESCA (LOW SAMPLE SIZE), STAD (LOW SAMPLE SIZE))
+	\end{frame}
+
+	\section{Differential Expression}
+	\begin{frame}{Finding Differentially Expressed Genes}
+		\begin{itemize}
+			\item Reads are count based hence they cannot be normally distributed
+		\end{itemize}
+	\end{frame}
 

### 2023-07-13 july18meeting_pres2/may30ccgmeeting.tex
Updates

@@
 	\section{Differential Expression}
-	\begin{frame}{Finding Differentially Expressed Genes}
+	%Our data do not fit the Poisson distribution
+	%(i.e. Poisson will underestimate variability leading to an increase in false positive DE genes).
+	\begin{frame}{Why does DESeq2 Use the Negative Binomial Distribution}
 		\begin{itemize}
 			\item Reads are count based hence they cannot be normally distributed
+			\item Variance tends to be greater than the mean, especially for genes with large mean expression values
+			\item  We need to account for this increase in variance using the Negative Binomial model 
 		\end{itemize}
+	\centering \includegraphics[width=5.5cm, height=5.5cm]{all_cancersmean_vs_variance_male.png}
+	\end{frame}
+
+	\begin{frame}{DESeq2 Accounts for Increased Variance}
+		\begin{enumerate}
+			\item Estimate gene-wise dispersions: captures biological variability in gene expression across samples
+			\item Fit negative binomial GLM to count data: This model incorporates the estimated gene-wise dispersions as a parameter to account for variability in the data
+			\item Use fitted GLM to identify differentially expressed genes \newline
+		\end{enumerate}
+		The use of linear models allows for more complex designs \newline \newline
+		$design = ~cancertype + condition$ \newline \newline
+		This tells DESeq2 to test the effect of condition while controlling for the effect of cancer type
+	\end{frame}
+
+	\begin{frame}{Finding Differentially Expressed Genes (DESeq2)}
+
 	\end{frame}

### 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
updates to ppt
updates

@@
 	\begin{frame}{Finding Differentially Expressed Genes (DESeq2)}
+		\includegraphics[width=12cm, height=8cm]{all_cancersvolcano_analysis_0.9.png
+		}
+	\end{frame}
 
+	\begin{frame}{Differences in Number of DE Genes Males vs Females}
+		\centering \includegraphics[width=9cm, height=5cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
+		\begin{itemize}
+			\item \textcolor{blue}{MALES: 358 up-regulated in tumor, 280 Down-regulated in tumor}
+			\item \textcolor{yellow}{FEMALES: 171 up-regulated in tumor, 70 down-regulated in tumor}
+		\end{itemize}
+	\end{frame}
+
+	\begin{frame}{Enriched Biological Processes Differ in Males and Females}
+		\includegraphics[width=12cm, height=8cm]{all_cancersgsea_exclusivedegenes_male_female_bp0.9.png}
+	\end{frame}
+
+	\begin{frame}
+		content...
 	\end{frame}

### 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
updates

@@
 
+	\begin{frame}{DE Genes Found in Males and Females are Enriched in Cancer-Related Processes}
+		\includegraphics[width=11cm, height=7cm]{all_cancersgsea_inbothdegenes_male_female_bp0.9.png}
+	\end{frame}
+
 	\begin{frame}{Enriched Biological Processes Differ in Males and Females}
@@
 
-	\begin{frame}
+	\section{Synthetic Lethality}
+	\begin{frame}{Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
+		content...
+	\end{frame}
+
+	\section{Supplementary Slides}
+	\begin{frame}{Supplementary 1 - Histogram Showing Normalization}
+		\includegraphics[width=12cm, height=7cm]{all_cancers-F_before_and_after_normalization_hist.png}
+	\end{frame}
+
+	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
 		content...

### 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
updates

@@
 	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
+		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1male.png}
+	\end{frame}
+
+	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
+		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1female.png}
+	\end{frame}
+
+	\begin{frame}
 		content...

### 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
Updates

@@
 	\section{Synthetic Lethality}
-	\begin{frame}{Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
-		content...
+	\begin{frame}{Next Steps: Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
+		\begin{enumerate}
+			\item Find all potential gene pair combinations using down regulated genes specific to males and specific to females \newline
+
+			\textcolor{blue}{${280 \choose 2 } = 39060$ Downregulated Male Pairs \newline}
+			\textcolor{yellow}{${70 \choose 2 } = 2415$ Downregulated Female Pairs}
+
+			\item Check whether the co-inactivation of a candidate SL pair is underrepresented in tumor samples
+		\end{enumerate}
 	\end{frame}
@@
 
-	\begin{frame}
-		content...
-	\end{frame}
-
 \end{document}

### 2023-07-20 sl_report_pmcrc/paper0904.tex
updates to report

@@
 	RNA sequencing (RNA-seq) data was obtained from The Cancer Genome Atlas (TCGA).
-Raw STAR aligned counts for tumor tissue and healthy tissue samples were utilized.
+Raw STAR (Spliced Transcripts Alignment to a Reference) aligned counts for tumor tissue and healthy tissue samples were collected.
 The Cancer Genome Atlas contains genomic information which spans 33 cancer types.
@@
 We first filtered out sex-biased cancers which include breast invasive carcinoma (BRCA), cervical cell carcinoma (CESC), ovarian serous cystadenocarcinoma (OV), prostate adenocarcinoma (PRAD), testicular germ cell tumors (TGCT), uterine corpus endometrial carcinoma (UCEC), and uterine carcinosarcoma (UCS).
+The reason for this was due to the fact that we are already aware of sex biases in these cancer types.
 Next, we filtered out blood cancers as well as cancers which lacked normal tissue gene expression samples.
 This included adrenocortical carcinoma (ACC), lymphoid neoplasm diffuse large b-cell lymphoma (DLBC), glioblastoma multiforme (GBM), acute myeloid leukemia (LAML), and brain lower grade glioma (LGG).
-Finally, we filtered out cancers that did not have any matching pairs of samples (NT and TP from same individual), as well as cancers that had extremely low numbers of matched sample pairs.
+The reason for this was due to the fact that we did not have adequate control samples to compare to tumor samples.
+Finally, we filtered out cancers that did not have any matching pairs of samples (NT and TP from same individual), as well as cancers that had less than 10 matched sample pairs across both males and females.
 This included mesothelioma (MESO), skin cutaneous melanoma (SKCM), thymoma (THYM), uveal melanoma (UVM), cholangiocarcinoma (CHOL), pancreatic adenocarcinoma (PAAD), pheochromocytoma and paraganglioma (PCPG), rectum adenocarcinoma (READ), and sarcoma (SARC).
+We selected matched tumor-normal sample pairs to help control for genetic background and other individual-specific factors that could influence gene expression.
+Once cancer types were selected, two pan-cancer raw count gene expression matrices were created, one for males and one for females.
 
 	\subsection{Pre-filtering and Normalization of Raw RNA-seq Count Data}
+
+	Raw count gene expression matrices for males and females were pre-filtered to remove genes unlikely to exhibit differential expression.
+For each matrix, we calculated the 90th quantile of overall gene expression as a threshold.
+For each gene in the matrix, we checked to see whether its expression was greater than the threshold in at least 1 sample, and removed genes where no samples showed expression greater than the quantile threshold.
 

### 2023-07-20 litrerview_notes_syntheticlethality/mutations_2.cpp
updates

@@ -1,170 +1,0 @@
-#include "functions.cpp"
-#include "getindex.cpp"
-#include <bits/stdc++.h>
-#include <iostream>
-#include <vector>
-#include <string>
-#include <ctime>
-#include <algorithm> //this is to get min element stuff, cool library
-#define numAA 20
-using namespace std;
-
-
-Ran myran(time(NULL)); //We will use 21 as the random seed right now, used in Poissondev too
-
-////////////////////////////////////////////////////////////////
-// FIRST FUNCTION TAKES AN INTEGER VALUE AND GENERATES A RANDOM/
-// AMINO ACID SEQUENCE OF THAT LENGTH                          / 
-////////////////////////////////////////////////////////////////
-
-std::string createSeq(int n){
-
-    char aminoAcids[numAA] = { 'G', 'A', 'L', 'M', 'F', 'W', 'K', 'Q', 'E', 'S', 'P', 'V', 
-                                   'I', 'C', 'Y', 'H', 'R', 'N', 'D', 'T' };
-
-    std::string protein = "";
-    for (int i = 0; i < n; i++){
-        protein += aminoAcids[myran.int64() % numAA];} //this rand() % 20 means in the range 0-19
-    
-    //std::cout << protein << "\n" << "\n" ;
-    return protein;
-}
-
-
-/*THIS FUNCTION WILL MUTATE THE SIMULATED PROTEIN SEQUENCE
-BY CHOOSING A RANDOM EXPONENTIAL DEVIATE (WITH MEAN = MUTATION RATE)
-FOR EACH AMINO ACID IN THE SEQUENCE AND SUBSEQUENTLY SELECTING
-THE AMINO ACID WITH THE LOWEST NUMBER (QUICKEST TO MUTATE) AND
-MUTATING IT RANDOMLY, THIS IS DONE SUCCESSIVELY TO PRODUCE A 
-PROTEIN AND CREATE A VECTOR OF VALUES SIMILAR TO ABOVE (SEP 21)     
-SEP 26 - PROCESS CHANGE, THIS FUNCTION WILL BE USED FOR AMINO ACID EXPANSION
-UPDATE NOV 15 - WE ARE MAKING THIS NEW FUNCTION THAT ASSIGNS DEVIATES
-BASED ON BOTH MUTATION RATES AND INDEL RATES AND THEN OUT OF BOTH VECTORS
-WE WILL FIND THE LOWEST AND EITHER MUTATE IT OR DO AN INS/DEL DEPENDING
[…]

### 2023-07-20 scripts_and_executables_examples/adding_Blatmerges_230418.py

@@ -1,59 +1,0 @@
-#!/usr/bin/env python3
-
-import pandas as pd
-import glob
-import numpy as np
-from sys import argv
-import os
-import datetime
-from pytz import timezone
-
-print('\n**********************')
-print('START TIME:', datetime.datetime.now(timezone('EST')))
-
-
-# loading in files! --> We're not going to use the scores yet!! --> VERY DIFFERENT, WE'RE JUST COUNTING UP THE NUMBER SO FAR
-res = pd.read_csv(str(argv[1]), sep='\t', header=None)
-print(res.head())
-res.columns = ['matches', 'misMatches', 'repMatches', 'nCount', 'qNumInsert', 'qBaseInsert', 'tNumInsert', 'tBaseInsert', 'strand', 'qName', 'qSize', 'qStart', 'qEnd', 'tName', 'tSize', 'tStart', 'tEnd', 'blockCount', 'blockSizes','qStarts', 'tStarts']
-# sorting the blat responses
-sorted_res = res.sort_values(by=['matches'], ascending=False)
-sorted_res = sorted_res.drop_duplicates(subset='qName', keep="first")
-
-
-# recall, no score csv --> using the 2nd argument for the svs
-svs = pd.read_csv(str(argv[2]), comment='#', sep='\t')
-print('PRE ANNOTATION')
-print(svs.head())
-
-# BY CHROMOSOME!!
-chromo = str(argv[3])
-svs = svs[svs.CHROM == chromo]
-ogrow = len(svs)
-print('og number of rows:', ogrow)
-
-#adding a column for a merge..
-for idx, row in svs.iterrows():
-	svs.loc[idx, 'qName'] = ('POST_'+ row['CHROM']+ '_' +str(row['POS'])+ '_' + str(row['SVlen'])+ '_' + row['SV_Type'])
-
-merged2 = pd.merge(svs, sorted_res, how = 'outer', on = 'qName', indicator = True)
-
-for idx, row in merged2.iterrows():
-	if row['_merge'] == 'left_only':
-		merged2.loc[idx, 'Homology'] = False
-	elif row['_merge'] == 'both':
-		merged2.loc[idx, 'Homology'] = True
-
-newrow = len(merged2)
-print('new number of rows:', newrow)
-
-# checking if the rows match (they should)
-if ogrow == newrow:
-	print('good merge')
[…]

### 2023-07-20 scripts_and_executables_examples/execute_Blatmerges_230418.sh

@@ -1,41 +1,0 @@
-#!/bin/bash
-#SBATCH --nodes=1
-#SBATCH --mem=300000
-#SBATCH --time=08:00:00
-#SBATCH --mail-type=ALL
-#SBATCH --mail-user=nb.boev@mail.utoronto.ca
-#SBATCH --job-name=blatmerge
-#SBATCH --output=blatmerge.o
-#SBATCH --error=blatmerge.e
-#SBATCH --qos=privileged
-
-source $HOME/virtenv/bin/activate
-module load scipy-stack
-
-# starting with the X chromo = 2023/04/18
-python adding_Blatmerges_230418.py \
-/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrX.Blatresults_full.csv \
-/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/SVlen/variants_freeze4_sv_insdel_alt.vcf.SVlength.csv chrX \
->> ./sv/adding_Blatmerges_230418.py.txt
-echo "Counting for Chromo X"
-rm /home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrX.Blatresults_full.csv
-
-
-# Running on Y chromo
-python adding_Blatmerges_230418.py \
-/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrY.Blatresults_full.csv \
-/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/SVlen/variants_freeze4_sv_insdel_alt.vcf.SVlength.csv chrY \
->> ./sv/adding_Blatmerges_230418.py.txt
-echo "Counting for Chromo Y"
-rm /home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chrY.Blatresults_full.csv
-
-# runnning through autosomes
-for i in {1..22};
-do
-	python adding_Blatmerges_230418.py \
-/home/nboev/projects/def-sushant/nboev/preprocess/HGSVC2_v2_integrated_callset/sv/Blat/results/chr${i}.Blatresults_full.csv \
[…]

### 2023-07-25 sl_report_pmcrc/paper0904.tex
updates

@@
 \usepackage{xcolor}
+\usepackage{booktabs}
+\usepackage{tabulary}
 
@@
 
+	\begin{table}[ht]
+	\centering
+	\caption{List of 12 TCGA Cancer Types With Number of Matched Tumor-Normal Samples in Males and Females.}
+	% Second version of table, with booktabs.
+	\begin{tabulary}{\textwidth}{CCCC}\toprule
+		& \multicolumn{3}{c}{TCGA information}
+		\\\cmidrule(lr){2-4}
+		& Cancer Type  & Matched Female Samples & Matched Male Samples\\\midrule
+		& BLCA & 9 & 10 \\
+		& COAD & 21 & 20 \\
+		& ESCA & 5 & 8 \\
+		& HNSC & 14 & 29 \\
+		& KICH & 12 & 13 \\
+		& KIRC & 20 & 52 \\
+		& KIRP & 10 & 22 \\
+		& LIHC & 22 & 28 \\
+		& LUAD & 34 & 24 \\
+		& LUSC & 14 & 37 \\
+		& STAD & 10 & 23 \\
+		& THCA & 42 & 17 \\
+		& \textcolor{red}{TOTAL} & \textcolor{red}{213} & \textcolor{red}{283} \\\bottomrule
+	\end{tabulary}
+	\end{table}
+
 	\subsection{Pre-filtering and Normalization of Raw RNA-seq Count Data}
@@
 For each matrix, we calculated the 90th quantile of overall gene expression as a threshold.
-For each gene in the matrix, we checked to see whether its expression was greater than the threshold in at least 1 sample, and removed genes where no samples showed expression greater than the quantile threshold.
+For each gene in the matrix, we checked to see whether its expression was greater than the threshold in at least 1 sample.
+We removed genes where no samples showed an expression value greater than the quantile threshold.
 

### 2023-07-26 sl_report_pmcrc/paper0904.tex
updates

@@
 
+	The pre-filtered matrices were then normalized using the DESeq2 package in R \citep{love2014moderated}.
+RNA-seq data must be normalized in order to account for factors that prevent the direct comparison of expression measures.
+The DESeq2 R package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
+
 	\subsection{Data Quality Assessment (PCA, NPManova)}

### 2023-07-26 sl_report_pmcrc/paper0904.tex
updates

@@
 The DESeq2 R package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
+Both sequencing depth (\# of reads generated per sample) and RNA composition (differences in composition of RNA molecules in a sample) are factors accounted for by the DESeq2 package.
+Raw counts are divided by size factors determined for each sample by computing the median ratio of gene counts relative to the geometric gene calculated per gene \citep{love2014moderated}.
+\figref{fig:1} shows the effects of normalization.
+
+	\begin{figure}[!h]
+		\centering	
+		\includegraphics[width=8.5cm, height=6cm]{normalied_male.png}\hfill
+		\includegraphics[width=8.5cm, height=6cm]{normalied_female.png}
+		\caption{Boxplots highlighting the distribution of raw RNA-seq data (top row) vs normalized RNA-seq data (bottom row) in males (left) and females (right) across normal tissue and tumor tissue samples from 12 TCGA cancer types.}
+		\label{fig:1}
+	\end{figure}
 

### 2023-07-28 sl_report_pmcrc/paper0904.tex
updates

@@
 RNA-seq data must be normalized in order to account for factors that prevent the direct comparison of expression measures.
-The DESeq2 R package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
+The DESeq2 package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
 Both sequencing depth (\# of reads generated per sample) and RNA composition (differences in composition of RNA molecules in a sample) are factors accounted for by the DESeq2 package.
 Raw counts are divided by size factors determined for each sample by computing the median ratio of gene counts relative to the geometric gene calculated per gene \citep{love2014moderated}.
-\figref{fig:1} shows the effects of normalization.
+\figref{fig:1} summarizes the effects of normalization for both male and female RNA-seq data.
+The distribution of counts across samples becomes much more consistent, thus making the samples comparable.
 
@@
 		\includegraphics[width=8.5cm, height=6cm]{normalied_female.png}
-		\caption{Boxplots highlighting the distribution of raw RNA-seq data (top row) vs normalized RNA-seq data (bottom row) in males (left) and females (right) across normal tissue and tumor tissue samples from 12 TCGA cancer types.}
+		\caption{Boxplots highlighting the distribution of raw RNA-seq data (top row) vs normalized RNA-seq data (bottom row) in males (left) and females (right) across normal tissue and tumor tissue samples from 12 TCGA cancer types.
+Each colour represents a specific cancer type.}
 		\label{fig:1}
@@
 	\subsection{Data Quality Assessment (PCA, NPManova)}
+
+	Once expression matrices were normalized
 

### 2023-08-01 sl_report_pmcrc/paper0904.tex
updates

@@
 
-	Once expression matrices were normalized
+	Once expression matrices were normalized, we performed principle component analysis (PCA) on the normalized counts to gain insights into potential factors contributing to the overall variance.
+Gene expression data is complex due to large number of variables (genes) present.
+PCA is a technique used to reduce dimensionality by transforming the data to a new set of variables (principle components) that summarize features of the data \citep{yeung2001principal}.
+In the context of this study, PCA is useful because it can take expression information from many genes, and reduce it down to fewer dimensions, making the data easier to explore.
+Using PCA, we explored the effects of tissue condition (normal vs tumor), and cancer type on gene expression.
+
+	\begin{figure}[!h]
+		\centering	
+		\includegraphics[width=\textwidth,height=9cm]{all_cancerspca_analysis_condition_0.9.png}
+		\caption{PCA plot highlighting the effect of tissue condition (normal vs tumor) on gene expression in males (left) and females (right) across 12 cancer types.
+Normal tissue samples are shown in red, tumor tissue samples are shown in blue.}
+		\label{fig:2}
+	\end{figure}
+
+	\figref{fig:2} shows the effects of tissue condition (normal vs tumor) on gene expression across the 12 cancer types that were considered in this study.
+In both males and females, clusters based on tissue condition were present, however there was more variability in clusters of tumor tissue samples compared to clusters of normal tissue samples.
+\figref{fig:3} shows the effects of both tissue condition and cancer type on gene expression.
+
+	\begin{figure}[!h]
+		\centering	
+		\includegraphics[width=\textwidth,height=9cm]{all_cancerspca_analysis_condition_cancertype_0.9.png}
+		\caption{PCA plot highlighting the effect of tissue condition (normal vs tumor) and cancer type on gene expression in males (left) and females (right) across 12 cancer types.}
+		\label{fig:3}
+	\end{figure}
+
+	When we added cancer type as a factor, we obtained much clearer clusters in the PCA plot, suggesting that cancer type contributes to the variation in gene expression.
 
@@
 
-%	\subsection{One Sample T-Test for Probability of Acceptance}
+	\subsection{Fishers Exact Test For Synthetic Lethality}
 
@@
 \clearpage\newpage
+\section{Code \& Supplementary Materials}
+
+The following files listed are located on the Graham Cluster at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc}
+
+Files used to obtain and format data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/scripts}
+\begin{itemize}
+\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_female.R}
+
+This script utilizes the tcgabiolinks package in R to retrieve gene expression data (RNA-seq) for a specific set of TCGA barcodes.
+The TCGA barcodes used in this script came from females of a specific cancer type.
+For each cancer type, a raw gene expression matrix is produced with genes as rows, and samples as columns.
+DO ON LOGIN NODE, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+
+\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_female.R}
+
+This script utilizes the tcgabiolinks package in R to retrieve gene expression data (RNA-seq) for a specific set of TCGA barcodes.
+The TCGA barcodes used in this script came from males of a specific cancer type.
+For each cancer type, a raw gene expression matrix is produced with genes as rows, and samples as columns.
+
+\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_female.R}
+
+This script utilizes the tcgabiolinks package in R to retrieve clinical metadata for a specific set of TCGA barcodes.
+The TCGA barcodes used in this script came from females of a specific cancer type.
+For each cancer type, a SummarizedExperiment (R object) is produced and clinical information such as age at diagnosis and tissue of origin can be extracted.
+
+\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_male.R}
+
+This script utilizes the tcgabiolinks package in R to retrieve clinical metadata for a specific set of TCGA barcodes.
+The TCGA barcodes used in this script came from males of a specific cancer type.
+For each cancer type, a SummarizedExperiment (R object) is produced and clinical information such as age at diagnosis and tissue of origin can be extracted.
+
+\item\texttt{extract\_clinical\_data\_females.R}
+
+This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for females.
+
+\item\texttt{extract\_clinical\_data\_males.R}
+
+This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for males. \newline
+
+\item\texttt{extract\_both\_tp\_nt\_samples\_females.R}
+
+This script
+
+\item\texttt{extract\_both\_tp\_nt\_samples\_females.R}
+
+This script
+
+\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_females.R}
+
+This script takes the gene expression matrices (with matched NT-TP samples) for females of each cancer type and merges them to create a raw pan-cancer matrix.
+Rows are genes and columns are samples.
+Samples from every cancer type utilized in the study are present in this matrix.
+
+\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_males.R}
+
+This script takes the gene expression matrices (with matched NT-TP samples) for males of each cancer type and merges them to create a raw pan-cancer matrix.
+Rows are genes and columns are samples.
+Samples from every cancer type utilized in the study are present in this matrix.
+
+\end{itemize}
+
+Files used to process and visualize data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/scripts}
+
+\begin{itemize}
+	\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_female.R}
+
+	This script
+
+	\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_male.R}
+
+	This script
+
+	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_female.R}
+
+	This script
+
+	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_male.R}
+
+	This script
+
+	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_female.R}
+
+	This script
+
+	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_male.R}
+
+	This script
+
+	\item\texttt{dge\_analysis\_deseq2.R}
+
+	This script
+
+	\item\texttt{gsea\_analysis\_postdeseq.R}
+
+	This script
+
+	\item\texttt{npmanova.R}
+
+	This script
+
+	\item\texttt{npmanova.R}
+
+	This script
+
+\end{itemize}
+
+\clearpage\newpage
 \section{References}
@@
 \newpage
-
-%The following files included these appendices are located at \texttt{/home/alext/scratch/abcmcmc-thesis4c12}
 

### 2023-08-04 sl_report_pmcrc/paper0904.tex
updates

@@
 \clearpage\newpage
-\section{Code \& Supplementary Materials}
+\section{Code \& Supplementary Materials (Transition Document)}
 
-The following files listed are located on the Graham Cluster at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc}
+All of the following files listed are located on the Graham Cluster at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc}
 
-Files used to obtain and format data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/scripts}
-\begin{itemize}
-\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_female.R}
+Scripts used to obtain and format data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/scripts}
+
+Matching executables for these scripts can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/executions}
+
+NOTE: THE NUMBERED ORDER THESE SCRIPTS APPEAR IN MATCHES THE ORDER IN WHICH THEY WERE RUN
+
+\begin{enumerate}
+\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_female.R} - SCRIPT
 
@@
 For each cancer type, a raw gene expression matrix is produced with genes as rows, and samples as columns.
-DO ON LOGIN NODE, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+This is the first script run in the workflow in order to download the TCGA rna-seq data for the cancer types we were interested in.
 
-\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_female.R}
+DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+
+\item\texttt{tcgabiolinks\_rawcount\_rnaseq\_download\_male.R} - SCRIPT
 
@@
 For each cancer type, a raw gene expression matrix is produced with genes as rows, and samples as columns.
+This is the first script run in the workflow in order to download the TCGA rna-seq data for the cancer types we were interested in.
+
+DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 
@@
 
+DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+
 \item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_male.R}
@@
 
-\item\texttt{extract\_clinical\_data\_females.R}
+DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+\item\texttt{extract\_clinical\_data\_females.R} - SCRIPT\newline
+\texttt{extract\_clinical\_data\_females.sh} - EXECUTABLE
 
@@
 
-\item\texttt{extract\_clinical\_data\_males.R}
+\item\texttt{extract\_clinical\_data\_males.R}- SCRIPT\newline
+\texttt{extract\_clinical\_data\_males.sh} - EXECUTABLE
 
-This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for males. \newline
+This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for males.
 
-\item\texttt{extract\_both\_tp\_nt\_samples\_females.R}
+\item\texttt{extract\_both\_tp\_nt\_samples\_females.R} - SCRIPT \newline
+\texttt{extract\_both\_tp\_nt\_samples\_females.sh} - EXECUTABLE
 
@@
 
-\item\texttt{extract\_both\_tp\_nt\_samples\_females.R}
+\item\texttt{extract\_both\_tp\_nt\_samples\_males.R} - SCRIPT \newline
+\texttt{extract\_both\_tp\_nt\_samples\_males.sh} - EXECUTABLE
 
@@
 
-\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_females.R}
+\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_females.R} - SCRIPT \newline
+\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_females.sh} - EXECUTABLE
 
@@
 
-\end{itemize}
+\end{enumerate}
 
-Files used to process and visualize data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/scripts}
+Scripts used to process and visualize data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/scripts}
+
+Matching executables for these scripts can be found at
+\texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/executions}
 
@@
 	\item\texttt{gsea\_analysis\_postdeseq.R}
-
-	This script
-
-	\item\texttt{npmanova.R}
 

### 2023-08-07 aug08meeting_pres4/may30ccgmeeting.tex
updates

--- 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
@@
 % Title page details: 
-\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers \\~\\ Project Updates} 
+\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers \\~\\ Summer Wrap Up} 
 \author{Alexander Turco}
-\date{July 18, 2023}
+\date{August 8, 2023}
 \logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}
@@
 	% Outline frame - SLIDE 2
-	%\begin{frame}{Overview}
+	\begin{frame}{Overview}
 
-		%\begin{center}
-		%\begin{minipage}{6cm}
+		\begin{center}
+		\begin{minipage}{6cm}
 
-		 % 		\begin{block}{} \hyperlink{link1}{Background Information} \end{block}
-		  %		\begin{block}{} The Game Design Process \end{block}
-		  %		\begin{block}{} Collection of Student Feedback \end{block}
-		  %		\begin{block}{} Results \end{block}
-		  		%\begin{block}{} Conclusions \end{block}
-		  %		\begin{block}{} Future Work \end{block}
+		  		\begin{block}{} Overall Objective and Approach \end{block}
+		  		\begin{block}{} Synthetic Lethality Analysis Update \end{block}
+		  		\begin{block}{} Limitations of the Study \end{block}
+		  		\begin{block}{} Next Steps/Future Work \end{block}
+		  		\begin{block}{} Takeaways \& Tips for Future Undergrads \end{block}
 
-		%\end{minipage}
-		%\end{center}
+		\end{minipage}
+		\end{center}
 
-	%\end{frame}
+	\end{frame}
 
@@
 
-	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
-	\begin{frame}{Recall: The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
-		\label{link1}
+	\begin{frame}{Overall Objective}
+		\begin{itemize}
+			%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
+			\item Can we build sex-specific synthetic lethality networks for various cancer types? \newline
 
-		\begin{itemize}
-			\item Synthetic lethal interactions describe the relationship between two genes whose coupled inactivation, but not their individual inactivation, causes cell death or reduces cell viability \newline
-
-			\includegraphics[width=11cm, height=2.5cm]{slbasic.png}
-
-			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
+			\item More specifically, we are trying to elucidate the differences in synthetic lethal interactions between males and females using a network based approach.
 		\end{itemize}
-			\centering
-			\large \textbf{Synthetic = Combining}
-			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
-	\end{frame}
-
-	% SLIDE 4 - I do Not Know if I am including this slide
-	% The Graphic for this slide is an  
-	% Entropy, which is measured by Shannon's Entropy equation is a measure of compositional complexity which uses the proportion of residue(s)
-	% in a subsequence to measure the compositional state of that subseequence - A lower variety of residues = lower entropy 
-	%\begin{frame}{Shannon's Entropy - MAYBE }
-
-	%		$H = -L\sum p_i log_2(p_i)$
-
-	%\end{frame}
-
-	%Slide 6
-	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway.
-PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process.
-In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
-	\begin{frame}{Recall: Synthetic Lethal Interactions are Harnessed for Precision Oncology}
-		\begin{itemize} %(Olaparib, Niraparib, Rucaparib, Talazoparib)
-			\item \small Four FDA approved anti-cancer drugs  are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors that work via a synthetic lethal mechanism
-
-			\centering
-			\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}
-
-			\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
-		\end{itemize}
-
-	\end{frame}
-
-	%EXPLAIN THE PROCESS OF ISLE HERE
-	\begin{frame}{Recall: Building Pan-Cancer Synthetic Lethality Networks}
-		\centering
-		\includegraphics[width=13cm, height=7cm]{isle.png}
-		%\movie[width=10cm,height=7cm,showcontrols=true]{\includegraphics[width=10cm, height=7cm]{pompe_thumbnail.png}}{pompe_trimmed.mp4}
-		\footnotetext[2]{\tiny\cite{lee2018harnessing}}
-	\end{frame}
-
-	\begin{frame}{Recall: Sex Differences Add an Additional Layer of Complexity}
-		Human sex differences are mainly caused by;
-		\begin{enumerate}
-			\item Gonadal hormone secretions
-			\item Genes located on the sex chromosomes (X and Y) \newline
-		\end{enumerate}
-		This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
-	\end{frame}
-
-	\section{Approach}
-	\begin{frame}{The Objective}
-		\begin{center}
-			%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
-			Can we build sex-specific synthetic lethality networks for various cancer types? \newline
-
-			More specifically, we are trying to elucidate the differences in synthetic lethal interactions between males and females using a network based approach.
-		\end{center}
 	\end{frame}
@@
 
+	\section{Approach}
 	\begin{frame}{Overall Project Workflow}

### 2023-08-07 aug08meeting_pres4/may30ccgmeeting.tex
updates

@@
 	\begin{frame}{Overall Project Workflow}
-		\includegraphics[width=12cm, height=5.2cm]{img2.png}
-	\end{frame}
-
-	\section{Data Collection}
-	\begin{frame}{Selection of TCGA Cancer Types}
-			\begin{columns}
-				\begin{column}{0.45\textwidth}
-					\includegraphics[width=5cm, height=8cm]{img3.png}
-				\end{column}
-				\begin{column}{0.5\textwidth}
-					\begin{itemize}
-						\item There is a lack of healthy tissue RNA-seq samples in the TCGA database (12 cancers with more than 10 pairs in M and F)
-						\item Raw count pan-cancer matrices created with 12 remaining TCGA cancer types (BLCA, COAD, ESCA, HNSC, KICH, KIRC, KIRP, LIHC, LUAD, LUSC, STAD, THCA)
-					\end{itemize}
-				\end{column}
-			\end{columns}			
-	\end{frame}
-
-	\begin{frame}{Creating and Pre-Filtering Pan-Cancer Expression Matrices}
-		\begin{itemize}
-			\item For each gene, if expression is $>$ 90th quantile of overall expression in AT LEAST 1 sample, we keep it, otherwise we filter it out
-			\item We reduce the amount of genes with extremely low expression, thereby reducing noise and improving sensitivity to detect differentially expressed genes (genes with weak expression are more susceptible to technical noise arising from library size, library composition, etc)
-		\end{itemize}
-	\end{frame}
-
-	%Both samples are sequenced at the same depth - sample A has 5 genes expressed at same level, Sample B has the same genes as sample A plus two highly expressed genes that take up 50% of the reads
-	\section{Data Processing}
-	\begin{frame}{Normalization for RNA-seq Data Analysis}
-		\centering Required to identify genes that are differentially expressed due to some biological phenomena and not due to technical variation
-
-		\includegraphics[width=9cm,height=7cm]{img4.png}
-	\end{frame}
-
-	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
-		\includegraphics[width=12cm, height=8cm]{all_cancers-F_before_and_after_normalization_v.png}
-	\end{frame}
-
-	\section{Data Processing}
-	\begin{frame}{Normal and Tumor Tissue Differ ACROSS 12 Cancer Types}
-		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_0.9.png}
-
-		Across all 12 cancer types, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (NPManova p-val = 0.0001)
-	\end{frame}
-
-	\begin{frame}{Normal and Tumor Tissue Differ WITHIN each Cancer}
-		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_cancertype_0.9.png}
-
-		Within each cancer type, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (Acceptions: Males: ESCA (LOW SAMPLE SIZE), Females: BLCA, ESCA (LOW SAMPLE SIZE), STAD (LOW SAMPLE SIZE))
-	\end{frame}
-
-	\section{Differential Expression}
-	%Our data do not fit the Poisson distribution
-	%(i.e. Poisson will underestimate variability leading to an increase in false positive DE genes).
-	\begin{frame}{Why does DESeq2 Use the Negative Binomial Distribution}
-		\begin{itemize}
-			\item Reads are count based hence they cannot be normally distributed
-			\item Variance tends to be greater than the mean, especially for genes with large mean expression values
-			\item  We need to account for this increase in variance using the Negative Binomial model 
-		\end{itemize}
-	\centering \includegraphics[width=5.5cm, height=5.5cm]{all_cancersmean_vs_variance_male.png}
-	\end{frame}
-
-	\begin{frame}{DESeq2 Accounts for Increased Variance}
-		\begin{enumerate}
-			\item Estimate gene-wise dispersions: captures biological variability in gene expression across samples
-			\item Fit negative binomial GLM to count data: This model incorporates the estimated gene-wise dispersions as a parameter to account for variability in the data
-			\item Use fitted GLM to identify differentially expressed genes \newline
-		\end{enumerate}
-		The use of linear models allows for more complex designs \newline \newline
-		$design = ~cancertype + condition$ \newline \newline
-		This tells DESeq2 to test the effect of condition while controlling for the effect of cancer type
-	\end{frame}
-
-	\begin{frame}{Finding Differentially Expressed Genes (DESeq2)}
-		\includegraphics[width=12cm, height=8cm]{all_cancersvolcano_analysis_0.9.png
-		}
-	\end{frame}
-
-	\begin{frame}{Differences in Number of DE Genes Males vs Females}
-		\centering \includegraphics[width=9cm, height=5cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
-		\begin{itemize}
-			\item \textcolor{blue}{MALES: 358 up-regulated in tumor, 280 Down-regulated in tumor}
-			\item \textcolor{yellow}{FEMALES: 171 up-regulated in tumor, 70 down-regulated in tumor}
-		\end{itemize}
-	\end{frame}
-
-	\begin{frame}{DE Genes Found in Males and Females are Enriched in Cancer-Related Processes}
-		\includegraphics[width=11cm, height=7cm]{all_cancersgsea_inbothdegenes_male_female_bp0.9.png}
-	\end{frame}
-
-	\begin{frame}{Enriched Biological Processes Differ in Males and Females}
-		\includegraphics[width=12cm, height=8cm]{all_cancersgsea_exclusivedegenes_male_female_bp0.9.png}
+		\includegraphics[width=12cm, height=8cm]{project_workflow_slpmcrc_2.png}
 	\end{frame}
@@
 	\section{Synthetic Lethality}
-	\begin{frame}{Next Steps: Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
+	\begin{frame}{Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
+		\begin{center}
+			\includegraphics[width=7cm, height=4cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
+		\end{center}
 		\begin{enumerate}
-			\item Find all potential gene pair combinations using down regulated genes specific to males and specific to females \newline
+			\item Find all potential gene pair combinations using differentially expressed genes \newline	
+			\textcolor{blue}{${1289 \choose 2 } = 830,116$ Downregulated Male Pairs \newline}
+			\textcolor{yellow}{${892 \choose 2 } = 397,386$ Downregulated Female Pairs}
 
-			\textcolor{blue}{${280 \choose 2 } = 39060$ Downregulated Male Pairs \newline}
-			\textcolor{yellow}{${70 \choose 2 } = 2415$ Downregulated Female Pairs}
-
-			\item Check whether the co-inactivation of a candidate SL pair is underrepresented in tumor samples
+			\item Identify pairs of genes that are BOTH perturbed (under and overexpressed) more frequently than INDIVIDUALLY perturbed using Fishers Exact Test
 		\end{enumerate}
@@
 
-	\section{Supplementary Slides}
-	\begin{frame}{Supplementary 1 - Histogram Showing Normalization}
-		\includegraphics[width=12cm, height=7cm]{all_cancers-F_before_and_after_normalization_hist.png}
-	\end{frame}
+	\begin{frame}{Fishers Exact Test}
 
-	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
-		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1male.png}
-	\end{frame}
-
-	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
-		\centering \includegraphics[width=9cm, height=8cm]{all_cancerspairedboxplot_cancertype_0.9_lcal1female.png}
 	\end{frame}

### 2023-08-07 aug08meeting_pres4/may30ccgmeeting.tex
updates

@@
 		  		\begin{block}{} Synthetic Lethality Analysis Update \end{block}
-		  		\begin{block}{} Limitations of the Study \end{block}
 		  		\begin{block}{} Next Steps/Future Work \end{block}
+		  		\begin{block}{} Limitations of the study \end{block}
 		  		\begin{block}{} Takeaways \& Tips for Future Undergrads \end{block}
@@
 
+	\section{Overall Objective and Approach}
 	\begin{frame}{Overall Objective}
@@
 
-	\section{Approach}
 	\begin{frame}{Overall Project Workflow}
@@
 
-	\section{Synthetic Lethality}
+	\section{Synthetic Lethality Analysis Update}
 	\begin{frame}{Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
@@
 			\item Find all potential gene pair combinations using differentially expressed genes \newline	
-			\textcolor{blue}{${1289 \choose 2 } = 830,116$ Downregulated Male Pairs \newline}
-			\textcolor{yellow}{${892 \choose 2 } = 397,386$ Downregulated Female Pairs}
+			\textcolor{blue}{${1289 \choose 2 } = 830,116$ Potential Male Pairs \newline}
+			\textcolor{yellow}{${892 \choose 2 } = 397,386$ Potential Female Pairs}
 
@@
 	\begin{frame}{Fishers Exact Test}
+		Used to determine association between two categorical variables (overexpression and underexpression) when dealing with small sample sizes
+		\begin{center}
+		\includegraphics[width=10cm, height=7cm]{Contingency.png}
+		\end{center}
+	\end{frame}
 
+	\begin{frame}{Ranking Genes by Number of Interactions}
+		\includegraphics[width=12cm, height=5cm]{interactions.png}
+
+		\begin{itemize}
+		\item ISLE Method SL list Matches: Males = 14 pairs, Females = 11 pairs
+		\item Experimentally Validated SL list Matches: Males = 21 pairs, Females = 2 pairs
+		\end{itemize}
+	\end{frame}
+
+	\section{Next Steps/Future Work}
+	\begin{frame}{Next Steps for Synthetic Lethality Analysis}
+		\begin{enumerate}
+			\item Refine the current list of synthetic lethal pairs by performing survival analysis to identify SL pairs that contribute to improved patient survival.
+			\item Build male and female synthetic lethal interaction networks
+		\end{enumerate}
+	\end{frame}
+
+	\begin{frame}{}
+		content...
 	\end{frame}

### 2023-08-08 aug08meeting_pres4/may30ccgmeeting.tex
updates

@@
 		\end{enumerate}
+	\includegraphics[width=12cm, height=6cm]{network_interactions.png}
 	\end{frame}
 
-	\begin{frame}{}
+	\section{Limitations}
+	\begin{frame}{A Major Limitation}
+		In many cancers, normal tissues are not available in The Cancer Genome Atlas	
+		\begin{center}
+			\includegraphics[width=6cm, height=6cm]{lack.png}
+		\end{center}
+	\footnotetext[1]{\tiny Figure From  \cite{zeng2019selecting}}
+	\end{frame}
+
+	\section{Takeaways \& Tips}
+	\begin{frame}{Experience Gained (Technical Skills)}
+
+	\end{frame}
+
+	\begin{frame}{Experience Gained (Personal Skills)}
 		content...
@@
 
+	\begin{frame}{Major Takeaways for Future Undergrads}
+		\begin{itemize}
+			\item 
+		\end{itemize}
+	\end{frame}
+
+	\begin{frame}{Thank You}
+		To Sushant for taking a chance on me
+
+		To everyone in the lab for the constant support and guidance
+
+		11/10 would do again
+
+		\begin{center}
+			\includegraphics[width=7cm, height=5cm]{lab.jpg}
+		\end{center}
+	\end{frame}
+
 \end{document}

### 2023-08-08 sl_report_pmcrc/paper0904.tex
updates

@@
 
-	When we added cancer type as a factor, we obtained much clearer clusters in the PCA plot, suggesting that cancer type contributes to the variation in gene expression.
+	When we added cancer type as a factor, we obtained much clearer clusters in the PCA plot, suggesting that cancer type is a factor that contributes to the variation in gene expression.
 
@@
 
+	\subsection{Gene Set Enrichment Analysis with ClusterProfiler}
+
 	\subsection{Fishers Exact Test For Synthetic Lethality}
 
-\section{Results}
+%\section{Results}
 
@@
 
-\section{Discussion}
+%\section{Discussion}
 
@@
 
-All of the following files listed are located on the Graham Cluster at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc}
+\hl{All of the following files listed are located on the Graham Cluster at} \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc}
 
-Scripts used to obtain and format data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/scripts}
+\hl{Scripts used to obtain and format data can be found at} \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/scripts}
 
-Matching executables for these scripts can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/executions}
+\hl{Matching executables for these scripts can be found at} \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/raw\_data/executions}
 
@@
 
-DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 
@@
 
-DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 
-\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_female.R}
+\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_female.R} - SCRIPT
 
@@
 
-DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 
-\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_male.R}
+\item\texttt{tcgabiolinks\_rawcount\_clinical\_download\_male.R} - SCRIPT
 
@@
 
-DO ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 \item\texttt{extract\_clinical\_data\_females.R} - SCRIPT\newline
@@
 
-This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for females.
+This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for females in the form of a dataframe.
 
@@
 
-This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for males.
+This script extracts the clinical information of interest from the SummarizedExperiment (R object). We obtain barcode, tissue, and age information for males in the form of a dataframe.
 
@@
 
-This script
+This script finds all females within a TCGA cancer type who have BOTH normal and tumor samples.
+The script does not account for individuals with more than one match, meaning that the script will take all tumor and healthy samples pertaining to an individual, not just one pair.
+In some cases, individuals had two normal tissue samples.
+The output from this script is a gene expression matrix for each cancer type with only matched normal and tumor samples.
 
@@
 
-This script
+This script finds all males within a TCGA cancer type who have BOTH normal and tumor samples.
+The script does not account for individuals with more than one match, meaning that the script will take all tumor and healthy samples pertaining to an individual, not just one pair.
+In some cases, individuals had two normal tissue samples.
+The output from this script is a gene expression matrix for each cancer type with only matched normal and tumor samples.
 
@@
 
-This script takes the gene expression matrices (with matched NT-TP samples) for females of each cancer type and merges them to create a raw pan-cancer matrix.
+This script takes the gene expression matrices (with matched Normal-tumor tissue samples) for females of each cancer type and merges them together to create a raw pan-cancer matrix with all 12 cancer types.
 Rows are genes and columns are samples.
-Samples from every cancer type utilized in the study are present in this matrix.
 
@@
 
-This script takes the gene expression matrices (with matched NT-TP samples) for males of each cancer type and merges them to create a raw pan-cancer matrix.
+This script takes the gene expression matrices (with matched Normal-tumor tissue samples) for males of each cancer type and merges them together to create a raw pan-cancer matrix with all 12 cancer types.
 Rows are genes and columns are samples.
-Samples from every cancer type utilized in the study are present in this matrix.
 
-\end{enumerate}
+\hl{Scripts used to process and visualize data can be found at} \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/scripts}
 
-Scripts used to process and visualize data can be found at \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/scripts}
-
-Matching executables for these scripts can be found at
+\hl{Matching executables for these scripts can be found at}
 \texttt{$\sim$/projects/def-sushant/alextu/\newline slproject\_pmcrc/processed\_data/executions}
 
-\begin{itemize}
-	\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_female.R}
+\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_female.R} - SCRIPT\newline
+\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_female.sh} - EXECUTABLE
 
@@
 
-	\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_male.R}
+\item\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_male.R} - SCRIPT\newline
+\texttt{deseq2\_extract\_normalized\_counts\_allcancers\_male.sh} - EXECUTABLE
 
@@
 
-	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_female.R}
+\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_female.R} - SCRIPT\newline
+\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_female.sh} - EXECUTABLE
 
@@
 
-	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_male.R}
+\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_male.R} - SCRIPT \newline
+\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_allcancers\_male.sh} - EXECUTABLE
 
@@
 
-	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_female.R}
-
-	This script
-
-	\item\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_male.R}
+\item\texttt{pca\_analysis\_mergedcancers\_quantiles\_prefiltering.R} - SCRIPT\newline
+\texttt{pca\_analysis\_mergedcancers\_quantiles\_prefiltering.sh} - EXECUTABLE
 
@@
 
-\end{itemize}
+\end{enumerate}
 

### 2023-08-09 sl_report_pmcrc/paper0904.tex
updates

@@
 	When we added cancer type as a factor, we obtained much clearer clusters in the PCA plot, suggesting that cancer type is a factor that contributes to the variation in gene expression.
+We wanted to examine factors that contribute to this variance as these factors may influence which genes are found to be differentially expressed.
+
+	We also verified the statistical significance of the difference between normal and tumor tissue expression across all cancer types and within each cancer type using a non-parametric multivariate analysis of variance (NPMANOVA), also known as a permutational multivariate analysis of variance (PERMANOVA).
 
@@
 
-%\section{Results}
+\section{Next Steps/Future Avenues}
 
@@
 
-\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_males.R}
+\item\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_males.R} - SCRIPT \newline
+\texttt{merge\_rawcounts\_both\_tp\_nt\_samples\_males.R} - EXECUTABLE
 
@@
 
-This script
+This script runs the differential gene expression analysis for females (through DESeq2) using the pan-cancer dataframe created from the last script above.
+This entails normalization of the expression data as well as hypothesis testing for each gene to determine whether expression is significantly different between tumor and normal tissue.
+This script also gathers all the metadata for the samples included in the pan-cancer matrix (condition (NT vs TP), cancer type, tissue type, age).
+Lastly, the script can account for different quantile thresholds set by the user.
+We filtered out genes with expression values less than some quantile value of overall gene expression.
+We tried 90th quantile and 50th quantile.
+The output from this script is a DESeq object that can be used to obtain differential gene expression information (raw counts, normalized counts, hypothesis test results for gene expression including log2FoldChange and adjusted p-values).
 
@@
 
-This script
+This script runs the differential gene expression analysis for males (through DESeq2) using the pan-cancer dataframe created from the last script above.
+This entails normalization of the expression data as well as hypothesis testing for each gene to determine whether expression is significantly different between tumor and normal tissue.
+This script also gathers all the metadata for the samples included in the pan-cancer matrix (condition (NT vs TP), cancer type, tissue type, age).
+Lastly, the script can account for different quantile thresholds set by the user.
+We filtered out genes with expression values less than some quantile value of overall gene expression.
+We tried 90th quantile and 50th quantile.
+The output from this script is a DESeq object that can be used to obtain differential gene expression information (raw counts, normalized counts, hypothesis test results for gene expression including log2FoldChange and adjusted p-values).
 
@@
 
-This script
+This script uses the DESeq object produced in the last step for females to extract raw and normalized gene counts for the pan-cancer matrix.
+The script creates boxplots showing raw counts before normalization vs normalized and transformed counts.
+Used to view the effects of normalization.
+The user can specify the quantile threshold value they would like to use in the script.
 
@@
 
-This script
+This script uses the DESeq object produced in the last step for males to extract raw and normalized gene counts for the pan-cancer matrix.
+The script creates boxplots showing raw counts before normalization vs normalized and transformed counts.
+Used to view the effects of normalization.
+The user can specify the quantile threshold value they would like to use in the script.
 
@@
 
-This script
+This script runs principal component analysis on the normalized count data for males AND females to detect sources of variation in the gene expression data.
+The user can specify the quantile threshold value they would like to use in the script.
 
@@
 
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+
 \item\texttt{gsea\_analysis\_postdeseq.R}
@@
 This script
+
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 

### 2023-08-10 sl_report_pmcrc/paper0904.tex
updates

@@
 	We also verified the statistical significance of the difference between normal and tumor tissue expression across all cancer types and within each cancer type using a non-parametric multivariate analysis of variance (NPMANOVA), also known as a permutational multivariate analysis of variance (PERMANOVA).
+This is a nonparametric test used to compare multivariate distributions of several groups and the null hypothesis for this test is that the centroids of all groups are equal \citep{anderson2005permutational}.
+In the case of this study, a low p-value indicates a significant difference in the centroids between normal tissue and tumor tissue.
+We did not have time to produce PCA plots that showed the centroids, but it would be beneficial to produce these plots in the future.
+We also did not perform an analysis that could identify the variance among the clusters so we could not statistically test if there was more heterogeneity among tumor tissue samples compared to normal tissue samples.
 

### 2023-08-10 sl_report_pmcrc/paper0904.tex
updates

@@
 We also did not perform an analysis that could identify the variance among the clusters so we could not statistically test if there was more heterogeneity among tumor tissue samples compared to normal tissue samples.
+\newline
+
+	\begin{table}[!h]
+		\centering
+		\vspace{-0.2cm}\includegraphics[width=18cm, height=5cm]{allfemale_npmanova.png}
+		\caption{Results from one-way NPMANOVA comparing centroids of normal tissue and tumor tissue gene expression in females across all 12 cancer types considered in the study}
+		\label{table:2}
+	\end{table}
+
+	\begin{table}[!h]
+		\centering
+		\vspace{-0.2cm}\includegraphics[width=18cm, height=5cm]{allmale_npmanova.png}
+		\caption{Results from one-way NPMANOVA comparing centroids of normal tissue and tumor tissue gene expression in males across all 12 cancer types considered in the study}
+		\label{table:3}
+	\end{table}
+
+Table 2 and Table 3 show the results of the one-way NPMANOVA (for females and males respectively) comparing the centroids of tumor tissue and normal tissue expression across all 12 cancer types together.
+In both males and females, the small p-value indicates a significant difference in normal vs tumor tissue when considering all 12 cancer types collectively.
+
+	\begin{table}[!h]
+		\centering
+		\includegraphics[width=18cm, height=6cm]{eachfemale_npmanova.png}
+		\caption{Results from pairwise NPMANOVA comparing centroids of normal tissue and tumor tissue gene expression in females for each individual cancer type considered in the study (full table not shown here)}
+		\label{table:4}
+	\end{table}
+
+	\begin{table}[!h]
+		\centering
+		\vspace{-0.4cm}\includegraphics[width=18cm, height=7cm]{eachmale_npmanova.png}
+		\caption{Results from pairwise NPMANOVA comparing centroids of normal tissue and tumor tissue gene expression in males for each individual cancer type considered in the study (full table not shown here)}
+		\label{table:5}
+	\end{table}
+
+	Table 4 and Table 5 show the results of the pairwise NPMANOVA (for females and males respectively) comparing the centroids of tumor tissue and normal tissue expression within each of the 12 cancer types individually.
+Shown in these tables are insignificant results (pairwise comparisons with adjusted p-value > 0.05) as it is easier to show which comparisons were not considered to be statistically significant.
+Only 3 rows of each table are shown, not the full tables.
 
 	\subsection{Differential Gene Expression Analysis with DESeq2}
+
+	Based on the above analysis, we decided to control for cancer type as it was found to be an important source of variation that could potentially confound the comparison between normal and healthy tissue conditions.
 

### 2023-08-11 sl_report_pmcrc/paper0904.tex
updates

@@
 
-	The pre-filtered matrices were then normalized using the DESeq2 package in R \citep{love2014moderated}.
+	The pre-filtered matrices were then normalized using the \texttt{DESeq2} package in R \citep{love2014moderated}.
 RNA-seq data must be normalized in order to account for factors that prevent the direct comparison of expression measures.
-The DESeq2 package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
-Both sequencing depth (\# of reads generated per sample) and RNA composition (differences in composition of RNA molecules in a sample) are factors accounted for by the DESeq2 package.
+The \texttt{DESeq2} package employs a median of ratios normalization method to account for the inherent biases associated with RNA-seq data.
+Both sequencing depth (\# of reads generated per sample) and RNA composition (differences in composition of RNA molecules in a sample) are factors accounted for by the \texttt{DESeq2} package.
 Raw counts are divided by size factors determined for each sample by computing the median ratio of gene counts relative to the geometric gene calculated per gene \citep{love2014moderated}.
@@
 	Based on the above analysis, we decided to control for cancer type as it was found to be an important source of variation that could potentially confound the comparison between normal and healthy tissue conditions.
+We used \texttt{DESeq2} to identify differentially expressed genes \citep{love2014moderated}.
+\texttt{DESeq2} utilizes the negative binomial distribution (AKA gamma-poisson distribution) to model the expression count data.
+The reason for using the negative binomial distribution as opposed to using the Poisson distribution (another distribution used to model count data) is due to the negative binomial distributions ability to account for variance (second paramter called dispersion).
+Typically with RNA-seq data, what we see is that the mean does not equal the variance.
+\figref{fig:4} shows this relationship and what we see is that the variance increases with mean expression (mean does not equal variance, hence Poisson is a poor choice).
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=10cm, height=10cm]{all_cancersmean_vs_variance_male.png}
+		\caption{Scatter plot showing the mean of gene counts (x-axis) vs variance of gene counts (y-axis) in males for all 12 cancer types considered in the study. The red line shows mean vs mean.}
+		\label{fig:4}
+	\end{figure}
+
+	\texttt{DESeq2} calculates dispersion estimates which accounts for the inherent variability in RNA-seq count data.
+Once dispersion estimates are calculated for each gene, a negative binomial generalized linear model (GLM) is fit for each gene which incorporates parameters for both mean and dispersion.
+Once GLM's are fit for each gene, a statistical test called the Wald test is employed to find differentially expressed genes between conditions of interest.
+Since \texttt{DESeq2} uses a linear model, it gives us the ability to create more complex designs \citep{love2014moderated}.
+In our analysis, we incorporated cancer type as a factor in the GLM.
+The following formula below was used by DESeq2 to test the effect of condition (normal vs tumor tissue) while controlling for the effect of cancer type.
+
+	\begin{equation}
+		design = cancer\_type + condition
+	\end{equation}
+
+	Genes found to be significantly differentially expressed were visualized using a volcano plot shown in \figref{fig:5}.
+Each gene is represented by an individual dot and the colour of the dot represents the gene passing a specific significance threshold.
+Red dots indicate that the gene had a p-value < 10e-20, and a log2FoldChange > 1.5 or < -1.5 whereas green dots indicate that the gene ONLY had a log2FoldChange > 1.5 or < 1.5.
+The log2FoldChange is a metric that quantifies the magnitude of change in gene expression between normal and tumor tissue conditions.
+Dots on the left side of each plot, with a log2FoldChange < -1.5, indicate donwregulation/decreased expression of the gene in tumor tissue compared to normal tissue.
+Dots on the right side of each plot, with a log2FoldChange > 1.5 indicate upregulation/increased expression of the gene in tumour tissue compared to normal tissue.
+We set the p-value threshold high when making these volcano plots to minimize the number of genes shown on the plot for aesthetic purposes.
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=\textwidth, height=12cm]{all_cancersvolcano_analysis_0.9.png}
+		\caption{Volcano plot highlighting the genes found to be differentially expressed in tumor tissue compared to normal tissue in males (left) and females (right).
+A red dot indicates the gene had a p-value < 10e-20 and log2FoldChange > 1.5 or < -1.5.
+A green dot indicates the gene only had a log2FoldChange > 1.5 or < -1.5}
+		\label{fig:5}
+	\end{figure}
+
+	The total number of differentially expressed genes are shown below in \figref{fig:6}.
+We found 1289 male genes to be significantly differentially expressed and 892 female genes to be significantly differentially expressed.
+(adj p-value < 0.05 AND |log2FoldChange| > 1.5).
+We compared the differentially expressed genes found in males and females and discovered 651 shared genes, 638 genes found only differentially expressed in males, and 241 genes found only differentially expressed in females.
+We also determined the number of genes found upregulated and downregulated in tumor tissue.
+In males, of the 1289 total genes found, 497 were found downregulated in tumor tissue while 792 were found upregulated in tumor tissue.
+In females, of the 892 total genes found, 287 were found downregulated in tumor tissue and 605 were found upregulated in tumor tissue.
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=8cm, height=5cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
+		\caption{Venn diagram showing the similarities and differences between differentially expressed genes found in males and females}
+		\label{fig:6}
+	\end{figure}
 
@@
 \section{Next Steps/Future Avenues}
+\begin{itemize}
+	\item Build male and female synthetic lethal interaction networks using results from Fishers exact test.
+Use the genes as nodes and the odds ratios as edges to build the networks.
+(Packages: \texttt{NetworkX} Python, \texttt{pyvis} Python).
+Find differences in the male and female networks.
+	\item Perform survival analysis on the SL pairs found using Fishers exact test and see which pairs contribute to improved patient survival.
+This could provide us with more clinically relevant synthetic lethal interactions.
+	\item Incorporate other types of data. We only included RNA-seq data, what about copy number variation, methylation, etc... this could strengthen the validity of the study.
+	\item Normal tissue proxy: There is a lack of normal tissue samples in the TCGA database.
+This restricted the number of cancer types we used in the study, as well as the sample size of the cancer types we did use.
+Can we use a different database to find matched samples?
+Can we use a machine learning approach to generate normal tissue expression data?
+Emailed the author of this paper, interesting stuff in terms of generating normal tissue expression but the author never got back to me with her code.
+\citep{zeng2019selecting}
+	\item Do males have higher incidence of lung cancer because more of them are smokers, or is it because they are male.
+Can we try and better understand if other factors (like societal factors) cause sex differences, or is it truly due to genetic differences in sex?
+	\item Menopause?
+\end{itemize}
 

### 2023-08-11 sl_report_pmcrc/paper0904.tex
updates

@@
 
-\lhead{Synthetic Lethality} %This needs to change
+\lhead{Sex Specific Synthetic Lethality} %This needs to change
 \rhead{Alexander Turco}
@@
 
+	We used the \texttt{clusterProfiler} R package to perform gene set enrichment analysis (GSEA) in order to gain insight into the functional groups of genes enriched in certain biological pathways.
+The package depends on annotation data from GO.db (Gene Ontology) and KEGG.db (Kyoto Encyclopedia of Genes and Genomes) in order to calculate enrichment test for GO and KEGG pathways \citep{yu2012clusterprofiler}.
+Essentially the analysis determines whether members of a gene set (unordered collections of genes that are functionally related) are randomly distributed in a list of differentially expressed genes, or primarily found at the top (overexpressed) or bottom (underexpressed) when sorted based on log2FoldChange \citep{yu2012clusterprofiler}.
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=17cm, height=10cm]{all_cancersgsea_alldegenes_male_female_bp0.9.png}
+		\caption{Dotplot showing the results of GSEA performed on ALL differentially expressed genes (892 female, 1289 male) found in females (left) and males (right).
+This example shows 10 activated and 10 suppressed enriched gene sets.}
+		\label{fig:7}
+	\end{figure}
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=\textwidth, height=11cm]{all_cancersgsea_exclusivedegenes_male_female_bp0.9.png}
+		\caption{Dotplot showing the results of GSEA performed on EXCLUSIVELY differentially expressed genes (241 female, 638 male) found in females (left) and males (right).
+This example shows 10 activated and 10 suppressed enriched gene sets.}
+		\label{fig:8}
+	\end{figure}
+
 	\subsection{Fishers Exact Test For Synthetic Lethality}
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=\textwidth, height=8cm]{interactions.png}
+		\caption{Barplot showing the genes involved in the most number of potential synthetic lethal pairs found using Fishers exact test}
+		\label{fig:9}
+	\end{figure}
 
@@
 Can we try and better understand if other factors (like societal factors) cause sex differences, or is it truly due to genetic differences in sex?
-	\item Menopause?
+	\item Menopause? Age Differences? Differences in medications?
 \end{itemize}
@@
 
-This script
+This script performs a lot of tasks.
 
@@
 
-This script
+This script performs gene set enrichment analysis using \texttt{clusterProfiler}.
 
@@
 
-This script
+This script performs a non parametric multivariate analysis of variance (NPMANOVA) in order to determine if differences between
 

### 2023-08-11 sl_report_pmcrc/paper0904.tex
updates

@@
 
-\section{Abstract}
+\section{Graphical Workflow}
+
+\begin{center}
+	\includegraphics[width=\textwidth, height=13cm]{project_workflow_slpmcrc.png}
+\end{center}
 
@@
 
+	\figref{fig:7} and \figref{fig:8} show the results of gene set enrichment analysis performed on all differentially expressed genes in males and females (\figref{fig:7}), and genes exclusively differentially expressed in males and females (\figref{fig:8}).
+On the y-axis of these plots are gene ontology (GO) biological processes/gene sets, and the size of the dot provides information regarding the number of genes from our input list that are associated with the given gene set/process.
+On the X-axis of these plots is a measure called gene ratio.
+The gene ratio is the number of genes in our input gene list associated with a given GO term, divided by the total number of input genes in that pathway.
+There are also two sides to the plot, one side shows "activated" while the other shows "suppressed".
+A gene set/process is labeled as activated when the majority of genes within that set show increased expression in tumor tissue compared to normal tissue.
+Conversely, a gene set/process is labeled as suppressed when the majority of genes within that set show decreased expression in tumor tissue compared to normal tissue.
+
+	When looking at the most significantly enriched biological processes among all the differentially expressed genes we found in males and females (\figref{fig:7}), we find processes related to cancer development such as regulation of transcription, cell cycle processes, and chromosome organization.
+When looking at the most significantly enriched biological processes among differentially expressed genes found only in males and only in females, we find processes related to sex differences such as inflammatory response to antigenic stimulus in women, and leydig cell differentiation in men.
+These results show that even across cancers that are not specific to one sex, sex differences in gene expression exist, and could potentially be driving factors that contribute to the differences in cancer progression and treatment seen in men and women.
+
 	\begin{figure}[!h]
@@
 
+	To find potential synthetic lethal interactions between the genes
+
 	\begin{figure}[!h]
 		\centering
-		\includegraphics[width=\textwidth, height=8cm]{interactions.png}
+		\includegraphics[width=16cm, height=6cm]{interactions.png}
 		\caption{Barplot showing the genes involved in the most number of potential synthetic lethal pairs found using Fishers exact test}
@@
 
+\clearpage\newpage
 \section{Next Steps/Future Avenues}
@@
 	\item Menopause? Age Differences? Differences in medications?
+	\item Include the centroids and statistical information on PCA plots to show the results of the NPMANOVA better
 \end{itemize}

### 2023-08-14 sl_report_pmcrc/paper0904.tex
updates

@@
 \rhead{Alexander Turco}
-\title{\sc Building sex-specific synthetic lethality network in cancer - Temp Name}
+\title{\sc Double Trouble: Understanding Sex Differences in Synthetic Lethal Interactions in Human Cancers}
 \author{\sc Alexander Turco}
@@
 
-	To find potential synthetic lethal interactions between the genes
+	To find potential synthetic lethal interactions between the genes found to be differentially expressed, we employed a Fisher exact test.
+The reason for doing this was based off the paper written by \citet{lee2018harnessing}.
+In this paper, a hypergeometric test (same as one-sided Fishers exact test) was employed to check whether the co-inactivation of potential synthetic lethal pairs were underrepresented among tumors.
+This stems from the idea that synthetic lethal pairs are deleterious to tumors, so they are more likley to be under negative selection \citep{lee2018harnessing}.
+For the purpose of our study, we did something a bit different.
+We used a Fisher exact test to identify pairs of genes that were both perturbed (both underexpressed or both overexpressed) more frequently than individually perturbed (one overexpressed or one underexpressed).
+Those pairs found to be both perturbed significantly more frequently than individually perturbed were the pairs we considered to be potential synthetic lethal.
 
@@
 		\includegraphics[width=16cm, height=6cm]{interactions.png}
-		\caption{Barplot showing the genes involved in the most number of potential synthetic lethal pairs found using Fishers exact test}
+		\caption{Barplot showing the genes involved in the most number of potential synthetic lethal pairs found using Fishers exact test.
+All genes shown here have an adjusted p-value < 0.05 and odds ratio (OR) > 2}
 		\label{fig:9}
 	\end{figure}
+
+	\figref{fig:9} shows the genes found to be involved in the largest number of significant candidate synthetic lethal gene pairs that we found using the Fisher test.
+It gives us a good indication of which genes act as major hubs for synthetic lethal interactions in males and females.
 
@@
 
-This script runs principal component analysis on the normalized count data for males AND females to detect sources of variation in the gene expression data.
+This script runs principal component analysis on the normalized count data for males AND females (again using the DESeq object from scripts 11 and 12) to detect sources of variation in the gene expression data.
 The user can specify the quantile threshold value they would like to use in the script.
+The plots produced by this script show differences in gene expression based on normal vs tumor tissue condition as well as differences in gene expression based on tissue condition AND cancer type.
+
+\item\texttt{npmanova.R}
+
+This script performs a non parametric multivariate analysis of variance (NPMANOVA) in order to determine if statistically significant differences exist between normal and tumor tissue conditions.
+Think of this test as a way to sort of statistically verify the differences we saw in the PCA.
 
@@
 
-\item\texttt{npmanova.R}
+\item\texttt{sl\_analysis.ipynb}
 
-This script performs a non parametric multivariate analysis of variance (NPMANOVA) in order to determine if differences between
+Synthetic lethality analysis.
+
+RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
 

### 2023-08-14 sl_report_pmcrc/paper0904.tex
updates

@@
 		\centering
-		\includegraphics[width=16cm, height=6cm]{interactions.png}
+		\includegraphics[width=\textwidth, height=7cm]{interactions_or4.png}
 		\caption{Barplot showing the genes involved in the most number of potential synthetic lethal pairs found using Fishers exact test.
-All genes shown here have an adjusted p-value < 0.05 and odds ratio (OR) > 2}
+All genes shown here have an adjusted p-value < 0.05 and odds ratio (OR) > 4}
 		\label{fig:9}
@@
 It gives us a good indication of which genes act as major hubs for synthetic lethal interactions in males and females.
+Ideally, the next thing we want to do is visualize these plots as a network, with nodes representing major synthetic lethal hub genes and edges representing the odds ratio and direction of regulation (up or down) between potential synthetic lethal pairs.
+
+	We verified that our approach produced legitimate results by comparing the SL pairs we found to SL pairs found in the literature.
+All of the data we compared our pairs to can be found in the supplementary materials of the paper by \citet{cheng2021synthetic}.
+\figref{fig:10} summarizes the results of the comparisons.
+Isle results represent SL pairs found using the ISLE approach \citep{lee2018harnessing}, while experimentally validated results represent SL pairs that have been experimentally tested in a wet-lab setting.
+The experimentally validated pairs were collected from 18 different studies (shRNA, sgRNA, isogenic cell line mutations, etc).
+In males, 14 pairs we found were also found by the ISLE approach, and 21 pairs we found were also experimentally validated SL pairs.
+In females, 11 pairs we found were also found by the ISLE approach and only 2 pairs we found were also experimentally validated SL pairs.
+We have not looked into the differences in SL pairs found in men and women (a network approach would make it easier to potentially find these differences in SL pair function between males and females)
+
+	\begin{figure}[!h]
+		\centering
+		\includegraphics[width=\textwidth, height=13cm]{validating_syntheticlethal_pairs.png}
+		\caption{Showing which synthetic lethal gene pairs we found in males and females were also found validated in the literature}
+		\label{fig:10}
+	\end{figure}
 
@@
 Find differences in the male and female networks.
+Is the function of the genes involved in the SL pairs specific to male or female biology?
+What types of processes are the different SL pairs involved in in both men and women?
 	\item Perform survival analysis on the SL pairs found using Fishers exact test and see which pairs contribute to improved patient survival.
@@
 
-This script performs a lot of tasks.
+This script performs a lot of tasks related to differential gene expression.
+Since the differential expression analysis was performed in scripts 11 and 12, we use the results of those scripts here.
+Essentially what this script does is filter the results of the differential gene expression analysis.
+The script finds all the genes with an adjusted p-value less than .05 and a log fold change greater than 1.5 (threshold for overexpression) or less than -1.5 (threshold for underexpression).
+Once signficantly differentially expressed genes are found, the script annotates these genes and creates boxplots that show the distribution of the genes found across the different chromosomes.
+Finally, the script compares the differentially expressed genes found in males and females to see which genes are found both differentially expressed in males and females and which are found exclusively differentially expressed in either males or females.
 
@@
 
-This script performs gene set enrichment analysis using \texttt{clusterProfiler}.
+This script performs gene set enrichment analysis using \texttt{clusterProfiler}. Refer to this site to better understand the code written in this script.
+\url{http://yulab-smu.top/biomedical-knowledge-mining-book/enrichment-overview.html}.
+This script uses the results produced from the last script (17).
+It performs gene set enrichment analysis on the differentially expressed genes found in BOTH sexes and found EXCLUSIVELY in each of the sexes.
 
@@
 
-Synthetic lethality analysis.
+This jupyter notebook contains all the code used to find synthetic lethal pairs using our differentially expressed genes of interest.
 

### 2023-08-15 sl_report_pmcrc/paper0904.tex
updates

@@
 This jupyter notebook contains all the code used to find synthetic lethal pairs using our differentially expressed genes of interest.
+The files utilized in the synthetic lethality analysis (gene lists, count data, etc) are included in the directory with the jupyter notebook (user can just go through jupyter notebook and see which files are utilized).
+This script essentially quantifies gene expression values into 3 categories (high, med, low) based on quantiles of overall gene expression.
+We create contingency tables for each gene, which summarize the quantity of samples that have a particular gene pair with HIGH-HIGH, LOW-LOW, HIGH-LOW, or LOW-HIGH expression.
+The script then conducts a Fisher exact test, adjusts p values using Benjamani Hochberg, and outputs all pairs with adjusted p-value < 0.05 and odds ratio > 4.
+The script also produces bar plots showing the genes involved in the most synthetic lethal interactions, AND the script finds which synthetic lethal pairs were found to be validated in the literature.
 
 RUN ON LOGIN NODE/LOCAL COMPUTER, NOT ON COMPUTE NODE, COMPUTE NODE CANT DOWNLOAD
+
+\end{enumerate}
+
+SCRIPTS NOT USED IN WORKFLOW BUT STILL COOL AND POTENTIALLY USEFUL
+
+\begin{enumerate}
+	\item \texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_female.R} - SCRIPT \newline
+	\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_female.sh} - EXECUTION
+
+	This script produces boxplots of normalized counts for each individual cancer type for females.
+It shows the difference in raw vs normalized counts across samples on a cancer specific level as opposed to a pan-cancer level.
+
+	\item \texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_male.R} - SCRIPT \newline
+	\texttt{deseq2\_preprocessing\_boxplots\_quantile\_prefiltering\_male.sh} - EXECUTION
+
+	This script produces boxplots of normalized counts for each individual cancer type for males.
+It shows the difference in raw vs normalized counts across samples on a cancer specific level as opposed to a pan-cancer level.
+
+	\item \texttt{paired\_boxplots.R} - SCRIPT
+
+	This script produces boxplots that show the difference between tumor tissue and normal tissue gene expression on a cancer specific level.
+There are two bars for each cancer type, one representing normal tissue expression and one representing tumor tissue expression.
+Lines are drawn between paired samples and a t-test is conducted to show significant differences in normal vs tumor expression across each cancer type.
 

### 2024-01-22 summer_2023/aug08meeting_pres4/may30ccgmeeting.tex
new presentation started

--- 2023-08-08 aug08meeting_pres4/may30ccgmeeting.tex
@@
 	\begin{frame}{Experience Gained (Technical Skills)}
-
+		\includegraphics[width=\textwidth,height=8cm]{technical_skills_gained.png}
 	\end{frame}
@@
 	\begin{frame}{Experience Gained (Personal Skills)}
-		content...
+		\begin{figure}
+			\includegraphics[width=\textwidth,height=8cm]{soft_skills_gained.png}
+		\end{figure}
 	\end{frame}
@@
 	\begin{frame}{Major Takeaways for Future Undergrads}
-		\begin{itemize}
-			\item 
-		\end{itemize}
+		\begin{enumerate}
+			\item Documentation and Organization: Document daily tasks in a manner that is easy for you AND others to understand.
+This helps keep track of your progress and helps explain things you did to other individuals in a clear and concise way.
+			\item Think like a scientist: You are surrounded by brilliant minds, pay attention to how they come up with a solution when you go to them with a problem.
+Be creative and try to develop your own solutions, new ideas are always welcome.
+			\item Be Confident: Speak up, ask questions, get involved. You are here to learn, nobody will shame you for that. Sometimes things can go much faster when you just ask.
+		\end{enumerate}
 	\end{frame}

### 2024-01-22 jan30meeting_pres1/may30ccgmeeting.tex
update to jan30 pres

--- 2023-07-15 july18meeting_pres2/may30ccgmeeting.tex
@@
 \usepackage{multimedia}
-\usepackage{graphicx}
 % For the Flowchart
@@
 % Title page details: 
-\title[BEAP Dec 2022]{Double Trouble: Understanding Sex Differences in Synthetic Lethal interactions in Human Cancers \\~\\ Project Updates} 
+\title[BEAP Dec 2022]{Current Affairs: Utilizing Oxford Nanopore Sequencing Data to Detect Non-Canonical DNA Structures} 
 \author{Alexander Turco}
-\date{July 18, 2023}
+\date{\today}
 \logo{\includegraphics[height=0.5cm, width=3cm]{logo.png}}
@@
 	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
+	\section{Background}
+	\begin{frame}{A Brief History Lesson}
+		\begin{itemize}
+			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
+			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
+		\end{itemize}
+
+		\centering
+		\huge \textbf{Synthetic = Combining}
+
+		\footnotetext[1]{\tiny\cite{nijman2011synthetic}}
+
+	\end{frame}
 
 	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
-	\begin{frame}{Recall: The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
+	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
 		\label{link1}
@@
 			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
+			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
 		\end{itemize}
-			\centering
-			\large \textbf{Synthetic = Combining}
-			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
+
 	\end{frame}
@@
 
+	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
+	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}
+
+		\begin{block}{\textit{To gain insight into gene function}}
+			 \begin{itemize}
+			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
+			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
+			 \end{itemize}
+		\end{block} \pause
+
+		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
+			\begin{itemize}
+				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g.
+Olaparib was the first drug to work via a synthetic lethal mechanism)
+				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
+				%Talk about BRCA1 and PARP inhibitors for olaparib lol
+			\end{itemize}
+		\end{block}
+
+	\footnotetext[3]{\tiny\cite{nijman2011synthetic,shen2018synthetic}}
+
+	\end{frame}
+
 	%Slide 6
@@
 In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
-	\begin{frame}{Recall: Synthetic Lethal Interactions are Harnessed for Precision Oncology}
-		\begin{itemize} %(Olaparib, Niraparib, Rucaparib, Talazoparib)
-			\item \small Four FDA approved anti-cancer drugs  are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors that work via a synthetic lethal mechanism
-
+	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
+		\begin{itemize}
+			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
+		\end{itemize}
 		\centering
@@
 		\footnotetext[4]{\tiny Figure From  \cite{o2015targeting}}
-		\end{itemize}
+	\end{frame}
 
+	%SLIDE 6 
+	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}
+
+	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics.
+Some aspects of this phenomena are; \newline
+
+	\begin{enumerate}
+		\item Variation in lifetime cancer risk
+		\item Variation in cancer onset age
+		\item Variation in the genes driving the cancer across tissue types \newline
+	\end{enumerate}
+		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
+	\end{frame}
+
+	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
+			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
+			\begin{center}
+				\includegraphics[width=10cm, height=5.7cm]{csl.png}
+			\end{center}
+		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
+	\end{frame}
+
+	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
+	\section{Approaches}
+
+	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
+		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type
+
+		\centering
+		\includegraphics[width=8cm, height=6cm]{tcl1.png}
+		\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
+	\end{frame}
+
+		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
+		\begin{columns}
+			\begin{column}{0.5\textwidth}
+					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
+					\centering
+					Stem Cell Divisions
+			\end{column}
+			\begin{column}{0.5\textwidth}
+				\includegraphics[width=6cm, height=3cm]{tcl3.png}
+				\centering
+				\vspace{-0.5cm}Methylation
+			\end{column}
+		\end{columns}	
+		\centering	
+	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
+	\end{frame}
+
+	\begin{frame}{Where did the Data come From?}
+		A recently published reference set of genome-wide cSLs common to many cancer types was utilized
+		\begin{center}
+				\includegraphics[width=6cm,height=6cm]{corenet.png}
+		\end{center}
+	\footnotetext[2]{\tiny\cite{lee2018harnessing}}
 	\end{frame}
@@
 	%EXPLAIN THE PROCESS OF ISLE HERE
-	\begin{frame}{Recall: Building Pan-Cancer Synthetic Lethality Networks}
+	\begin{frame}{Building Pan-Cancer Synthetic Lethality Networks}
 		\centering
@@
 
-	\begin{frame}{Recall: Sex Differences Add an Additional Layer of Complexity}
+	%%% EXPLAIN NSSE SURVEY A BIT
+	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
+	\begin{frame}{There is Variability in Synthetic Lethal Interactions Depending on Cancer Type}
+
+		\textbf{\color{blue}Pan-Cancer Analysis:} Analysis of multiple types of cancer simultaneously, aims to identify commonalities shared across different types
+
+		\textbf{\color{blue}Cancer-Specific Analysis:} Analysis of individual cancer types to better understand the variation within a specific cancer type \vspace{1cm}
+
+		\begin{itemize}
+			\item Pan-cancer analyses may overlook cancer-specific differences
+			\item Availability of datasets for all cancer types varies, this may limit statistical power of pan-cancer analyses
+		\end{itemize}
+
+	\end{frame}
+
+	%%%%RESULTS BEGIN HERE
+	\section{Results}
+
+	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
 		Human sex differences are mainly caused by;
@@
 
-	\section{Approach}
+	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
+		\centering
+		\includegraphics[width=6cm,height=7cm]{sex.png}
+		\footnotetext[6]{\tiny\cite{oliva2020impact}}
+	\end{frame}
+
 	\begin{frame}{The Objective}
@@
 
-	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
-
-	\begin{frame}{Overall Project Workflow}
-		\includegraphics[width=12cm, height=5.2cm]{img2.png}
-	\end{frame}
-
-	\section{Data Collection}
-	\begin{frame}{Selection of TCGA Cancer Types}
-			\begin{columns}
-				\begin{column}{0.45\textwidth}
-					\includegraphics[width=5cm, height=8cm]{img3.png}
-				\end{column}
-				\begin{column}{0.5\textwidth}
-					\begin{itemize}
-						\item There is a lack of healthy tissue RNA-seq samples in the TCGA database (12 cancers with more than 10 pairs in M and F)
-						\item Raw count pan-cancer matrices created with 12 remaining TCGA cancer types (BLCA, COAD, ESCA, HNSC, KICH, KIRC, KIRP, LIHC, LUAD, LUSC, STAD, THCA)
-					\end{itemize}
-				\end{column}
-			\end{columns}			
-	\end{frame}
-
-	\begin{frame}{Creating and Pre-Filtering Pan-Cancer Expression Matrices}
-		\begin{itemize}
-			\item For each gene, if expression is $>$ 90th quantile of overall expression in AT LEAST 1 sample, we keep it, otherwise we filter it out
-			\item We reduce the amount of genes with extremely low expression, thereby reducing noise and improving sensitivity to detect differentially expressed genes (genes with weak expression are more susceptible to technical noise arising from library size, library composition, etc)
-		\end{itemize}
-	\end{frame}
-
-	%Both samples are sequenced at the same depth - sample A has 5 genes expressed at same level, Sample B has the same genes as sample A plus two highly expressed genes that take up 50% of the reads
-	\section{Data Processing}
-	\begin{frame}{Normalization for RNA-seq Data Analysis}
-		\centering Required to identify genes that are differentially expressed due to some biological phenomena and not due to technical variation
-
-		\includegraphics[width=9cm,height=7cm]{img4.png}
-	\end{frame}
-
-	\begin{frame}{Normalizing Pan-Cancer Expression Matrices}
-		\includegraphics[width=12cm, height=8cm]{all_cancers-F_before_and_after_normalization_v.png}
-	\end{frame}
-
-	\section{Data Processing}
-	\begin{frame}{Normal and Tumor Tissue Differ ACROSS 12 Cancer Types}
-		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_0.9.png}
-
-		Across all 12 cancer types, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (NPManova p-val = 0.0001)
-	\end{frame}
-
-	\begin{frame}{Normal and Tumor Tissue Differ WITHIN each Cancer}
-		\includegraphics[width=12cm, height=6cm]{all_cancerspca_analysis_condition_cancertype_0.9.png}
-
-		Within each cancer type, the observed variation between healthy and tumor tissue samples is unlikely to have occurred by chance alone (Acceptions: Males: ESCA (LOW SAMPLE SIZE), Females: BLCA, ESCA (LOW SAMPLE SIZE), STAD (LOW SAMPLE SIZE))
-	\end{frame}
-
-	\section{Differential Expression}
-	%Our data do not fit the Poisson distribution
-	%(i.e. Poisson will underestimate variability leading to an increase in false positive DE genes).
-	\begin{frame}{Why does DESeq2 Use the Negative Binomial Distribution}
-		\begin{itemize}
-			\item Reads are count based hence they cannot be normally distributed
-			\item Variance tends to be greater than the mean, especially for genes with large mean expression values
-			\item  We need to account for this increase in variance using the Negative Binomial model 
-		\end{itemize}
-	\centering \includegraphics[width=5.5cm, height=5.5cm]{all_cancersmean_vs_variance_male.png}
-	\end{frame}
-
-	\begin{frame}{DESeq2 Accounts for Increased Variance}
-		\begin{enumerate}
-			\item Estimate gene-wise dispersions: captures biological variability in gene expression across samples
-			\item Fit negative binomial GLM to count data: This model incorporates the estimated gene-wise dispersions as a parameter to account for variability in the data
-			\item Use fitted GLM to identify differentially expressed genes \newline
-		\end{enumerate}
-		The use of linear models allows for more complex designs \newline \newline
-		$design = ~cancertype + condition$ \newline \newline
-		This tells DESeq2 to test the effect of condition while controlling for the effect of cancer type
-	\end{frame}
-
-	\begin{frame}{Finding Differentially Expressed Genes (DESeq2)}
-		\includegraphics[width=12cm, height=8cm]{all_cancersvolcano_analysis_0.9.png
-		}
-	\end{frame}
-
-	\begin{frame}{Differences in Number of DE Genes Males vs Females}
-		\centering \includegraphics[width=9cm, height=5cm]{all_cancersvenndiagram_malede_vs_femalede_0.9.png}
-		\begin{itemize}
-			\item \textcolor{blue}{MALES: 358 up-regulated in tumor, 280 Down-regulated in tumor}
-			\item \textcolor{yellow}{FEMALES: 171 up-regulated in tumor, 70 down-regulated in tumor}
-		\end{itemize}
-	\end{frame}
-
-	\begin{frame}{DE Genes Found in Males and Females are Enriched in Cancer-Related Processes}
-		\includegraphics[width=11cm, height=7cm]{all_cancersgsea_inbothdegenes_male_female_bp0.9.png}
-	\end{frame}
-
-	\begin{frame}{Enriched Biological Processes Differ in Males and Females}
-		\includegraphics[width=12cm, height=8cm]{all_cancersgsea_exclusivedegenes_male_female_bp0.9.png}
-	\end{frame}
-
-	\section{Synthetic Lethality}
-	\begin{frame}{Using DE Genes to Find Sex-Specific Synthetic Lethal Pairs}
-		content...
-	\end{frame}
-
-	\section{Supplementary Slides}
-	\begin{frame}{Supplementary 1 - Histogram Showing Normalization}
-		\includegraphics[width=12cm, height=7cm]{all_cancers-F_before_and_after_normalization_hist.png}
-	\end{frame}
-
-	\begin{frame}{Supplementary 2 - Gene Expression Across Tissues}
-		content...
-	\end{frame}
-
 \end{document}

### 2024-01-26 jan30meeting_pres1/may30ccgmeeting.tex
updates

@@
 	\section{Background}
-	\begin{frame}{A Brief History Lesson}
+	\begin{frame}{The Structure of DNA: A Brief History Lesson}
 		\begin{itemize}
-			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
-			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
+			\item In 1953, Watson, Crick, Wilkins, Franklin, and Gosling were the first to describe the structure of DNA \newline
+			\item They discovered the \textcolor{blue}{right-handed double helix} (canonical B-form DNA), the most common form found in cells \newline
 		\end{itemize}
@@
 		\centering
-		\huge \textbf{Synthetic = Combining}
+		\includegraphics[width=11cm, height=3.5cm]{rosy.png}
 
@@
 	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
-	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
+	\begin{frame}{DNA Can Adopt Alternative Structures}
 		\label{link1}
@@
 		\begin{itemize}
-			\item Synthetic lethal interactions describe the relationship between two genes whose coupled inactivation, but not their individual inactivation, causes cell death or reduces cell viability \newline
+			\item Now, more than 15 types of DNA structure that differ from the canonical B-form have been reported (non-canonical or non-B form DNA) \newline
+			\item Through sequencing of the human genome, we now know over half the genome is composed of repetitive elements - these were initially thought to be `junk DNA' \newline
 
-			\includegraphics[width=11cm, height=2.5cm]{slbasic.png}
+			\begin{center}
+				\includegraphics[width=9cm, height=1.2cm]{repeats.png}
+			\end{center}
 
-			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
+			\item A crucial feature of some repetitive sequences is their ability to fold into non-canonical DNA structures (non-B DNA)
 			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
@@
 	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
-	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}
+	\begin{frame}{Types of Non-canonical DNA structures}
 
-		\begin{block}{\textit{To gain insight into gene function}}
-			 \begin{itemize}
-			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
-			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
-			 \end{itemize}
-		\end{block} \pause
-
-		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
-			\begin{itemize}
-				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g.
-Olaparib was the first drug to work via a synthetic lethal mechanism)
-				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
-				%Talk about BRCA1 and PARP inhibitors for olaparib lol
-			\end{itemize}
-		\end{block}
+		\centering
+		\includegraphics[width=9cm, height=6.5cm]{nonb_types.png}
 
@@
 In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
-	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
+	\begin{frame}{Non-Canonical DNA Structures are Involved in Biological Processes}
+
+		Non-B DNA structures have been shown to co-localize with \textcolor{blue}{functional genomic loci} (promoters, enhancers, etc) and \textcolor{blue}{genetic instability hotspots} \newline
+
+		This suggests a role for non-B DNA in vital cellular events such as;
 		\begin{itemize}
-			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
+			\item Regulation of transcription
+			\item Regulation of DNA replication and recombination
+			\item Regulating genome integrity
 		\end{itemize}
+
 		\centering
-		\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}
+		%\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}
 
@@
 	%SLIDE 6 
-	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}
+	\begin{frame}{Diseases Associated with Non-Canonical DNA structures}
 
-	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics.
-Some aspects of this phenomena are; \newline
+	\textbf{\color{blue}Repeat Expansion Diseases:} Expansions of non-B DNA structure-forming repeats have been implicated in many neurodegenerative and neuromuscular diseases. \newline
 
-	\begin{enumerate}
-		\item Variation in lifetime cancer risk
-		\item Variation in cancer onset age
-		\item Variation in the genes driving the cancer across tissue types \newline
-	\end{enumerate}
-		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
+	\textbf{\color{blue}Genetic Instability Diseases:} Non-canonical DNA structures are associated with increased mutability (point mutations, deletions, insertions and chromosomal translocations)
+
+	\begin{itemize}
+		\item Enriched at chromosomal breakpoints in translocation-related cancers such as lymphomas and leukaemias.
+		\item Can be recognized by DNA repair proteins, triggering error-generating repair processes
+		\item G-quadruplexes are present within most human oncogenic promoters and at telomeres - a current theraputic target to downregulate transcription or block telomere elongation in cancer cells.
+	\end{itemize}
+
 	\end{frame}
 
-	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
-			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
+	\begin{frame}{How are Non-B Structures Detected in the Genome?}
 			\begin{center}
-				\includegraphics[width=10cm, height=5.7cm]{csl.png}
+				\vspace{-1cm}
+				\includegraphics[width=12cm, height=8cm]{Predicting_nonb_structures.png}
+				\vspace{-1.5cm}
 			\end{center}
-		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
+		These approaches are based primarily on \textcolor{blue}{DNA sequence motifs}, which are \textcolor{blue}{necessary}, but \textcolor{blue}{insufficient} for formation and are not available for all non-B DNA structures
+
 	\end{frame}
@@
 
-	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
+	\begin{frame}{A Promising New Approach}
 		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type
@@
 
-		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
-		\begin{columns}
-			\begin{column}{0.5\textwidth}
-					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
+	\begin{frame}{Oxford Nanopore Sequencing Technology}
 		\centering
-					Stem Cell Divisions
-			\end{column}
-			\begin{column}{0.5\textwidth}
-				\includegraphics[width=6cm, height=3cm]{tcl3.png}
+		\begin{figure}[!htb]
 			\centering
-				\vspace{-0.5cm}Methylation
-			\end{column}
-		\end{columns}	
+			\begin{minipage}{.3\textwidth}
 				\centering
+				\includegraphics[width=3.5cm, height=6cm]{ont_machine.jpg}
+				\caption{ONT Sequencer}
+				\label{fig:prob1_6_2}
+			\end{minipage}%
+			\begin{minipage}{0.7\textwidth}
+				\centering
+				\movie[width=8cm, height=4.5cm]{placeholder box}{ont.mp4}
+				\caption{Inside the Nanopore}
+				\label{fig:prob1_6_1}
+			\end{minipage}
+		\end{figure}
+
 	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}

### 2024-01-26 jan30meeting_pres1/may30ccgmeeting.tex
updates

@@
 	%%%%% APPRACH SECTION BEGINS HERE, 2 SLIDES
-	\section{Approaches}
 
-	\begin{frame}{A Promising New Approach}
-		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type
+	\begin{frame}{Third Generation Sequencing: A Promising New Approach}
+		\textbf{\color{blue}Single Molecule, Real Time Sequencing (SMRT):} Pacbio's third generation sequencing machine
+		\begin{itemize}
+			\item Emits a fluorescent pulse when nucleotide is detected - the time interval between two pulses is called the interpulse duration (IPD)
+			\item Guiblet et al (2018), showed that there is a significant divergence between IPDs in non-B DNA motif regions compared to B-DNA regions
+		\end{itemize}
 
 		\centering
-		\includegraphics[width=8cm, height=6cm]{tcl1.png}
-		\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
+		\includegraphics[width=10cm, height=4.5cm]{sequencing_evolution.png}
 	\end{frame}
@@
 
-	\begin{frame}{Where did the Data come From?}
+	\begin{frame}{Predicting Non-B Structures From Nanopore Sequencing}
 		A recently published reference set of genome-wide cSLs common to many cancer types was utilized

### 2024-01-27 jan30meeting_pres1/may30ccgmeeting.tex
updates

@@
 	\begin{frame}{Predicting Non-B Structures From Nanopore Sequencing}
-		A recently published reference set of genome-wide cSLs common to many cancer types was utilized
-		\begin{center}
-				\includegraphics[width=6cm,height=6cm]{corenet.png}
-		\end{center}
+		A recently published paper utilized translocation times from ONT sequencing to predict non-B DNA structures (citation)
+		\begin{itemize}
+			\item Developed the first computational pipeline and a novel unsupervised deep statistical model for predicting non-B DNA structures \newline
+		\end{itemize}
+
+		Benefits  of unsupervised approach;
+		\begin{enumerate}
+			\item non-B database labels are noisy (just because motif is present does not mean structure is)
+			\item Even if high quality labeling for non-B DNA were available, substantially more B-DNA samples are available
+			\item Unknown non-B structures or non-B DNA without sequence motifs cannot be modelled by a supervised approach
+		\end{enumerate}
 	\footnotetext[2]{\tiny\cite{lee2018harnessing}}
@@
 	%EXPLAIN THE PROCESS OF ISLE HERE
-	\begin{frame}{Building Pan-Cancer Synthetic Lethality Networks}
-		\centering
-		\includegraphics[width=13cm, height=7cm]{isle.png}
+	\begin{frame}{GoFAE-DND: Deep Statistical modelling of non-B DNA}
+		Anomaly Detection Problem: Identifying patterns within data that deviate significantly from the norm or expected behaviour of the majority of the data
+		\begin{center}
+		\includegraphics[width=12cm, height=5cm]{gofae.jpeg}
+		\end{center}
 		%\movie[width=10cm,height=7cm,showcontrols=true]{\includegraphics[width=10cm, height=7cm]{pompe_thumbnail.png}}{pompe_trimmed.mp4}
@@
 	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
-	\begin{frame}{There is Variability in Synthetic Lethal Interactions Depending on Cancer Type}
+	\begin{frame}{Results of the Model}
 
@@
 
-	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
+	\begin{frame}{The Objective}
 		Human sex differences are mainly caused by;
@@
 
-	\begin{frame}{Sex Differences Add an Additional Layer of Complexity}
+	\begin{frame}{Preprocessing Workflow}
 		\centering
@@
 
-	\begin{frame}{The Objective}
-		\begin{center}
-		%\includegraphics[width=11cm, height=6cm]{types_of_courses_taken.jpg}
-		Can we build sex-specific synthetic lethality networks for various cancer types? \newline
-
-		More specifically, we are trying to elucidate the differences in synthetic lethal interactions between males and females using a network based approach.
-		\end{center}
-	\end{frame}
-
 \end{document}

### 2024-01-29 jan30meeting_pres1/may30ccgmeeting.tex
updates

@@
 	%%% WHEN YOU TALK MENTION THAT THEY DID THIS SURVEY AFTER PLAYING THE GAME IN CLASS
-	\begin{frame}{Results of the Model}
-
-		\textbf{\color{blue}Pan-Cancer Analysis:} Analysis of multiple types of cancer simultaneously, aims to identify commonalities shared across different types
-
-		\textbf{\color{blue}Cancer-Specific Analysis:} Analysis of individual cancer types to better understand the variation within a specific cancer type \vspace{1cm}
-
-		\begin{itemize}
-			\item Pan-cancer analyses may overlook cancer-specific differences
-			\item Availability of datasets for all cancer types varies, this may limit statistical power of pan-cancer analyses
-		\end{itemize}
+	\begin{frame}{Model Performance}
+		At an FDR control level $\alpha = 0.2$, SVM and GoFAE-DND generated the most novelties, with GoFAE-DND yielding the most predictions for all non-B types besides G4
+		\begin{center}
+			\vspace{-1cm}
+			\includegraphics[width=12cm, height=8cm]{model_performance_gofae.png}
+		\end{center}
 
@@
 	\begin{frame}{The Objective}
-		Human sex differences are mainly caused by;
+		Given the dramatic increase in genome-scale data produced using ONT platforms, and the relevance of non-B structures in human cancers; \newline
 		\begin{enumerate}
-			\item Gonadal hormone secretions
-			\item Genes located on the sex chromosomes (X and Y) \newline
+			\item Utilize the model to analyze nanopore samples from the human pangenome reference consortium (HPRC)
+			\item Eventually improve the model, with an emphasis on the detection of G4 quadruplexes - Linking methylation, gene expression profiles which are available for HPRC samples\newline
 		\end{enumerate}
-	This leads to differences in the frequency of certain cancer types and the efficacy of treatments in males and females
 	\end{frame}
